---
title: "01_ComputeQCMetrics"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  
  #path to project
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"

  # How to load input objects
  use_seurat_list: true                    # TRUE: read list RDS; FALSE: load individual RDS using sample table
  seurat_list_rds: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/seurat_objects_by_sample.rds"
  
  # If use_seurat_list == FALSE, use these to load objects from disk
  sample_table: "" # Same table used in 00_CreateSeuratFromCounts
  sample_id_col: ""
  seurat_objects_dir: "" # Directory containing <sample_id>.rds files
  
  # QC parameters
  mito_genes: !expr character(0)
  qc_proxy_vars: !expr c("percent.mito", "nCount_RNA", "nFeature_RNA")
  nmad_lower: !expr c(Inf, 2, 2)
  nmad_upper: !expr c(2, 2, 2)

#  qc_proxy_vars: !r c("percent.mito", "nCount_RNA", "nFeature_RNA")
#  nmad_lower:    !r c(Inf, 2, 2)          # lower nmads for qc_proxy_vars (percent.mito, nCount_RNA, nFeature_RNA)
#  nmad_upper:    !r c(2, 2, 2)            # upper nmads for qc_proxy_vars
  qc_log: FALSE

  # Doublet detection
  run_doublet_detection: FALSE
  doublet_diet_seurat: TRUE                # if TRUE, remove SCT assay etc after DoubletFinder

  # Merged output
  merged_raw_rds: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/merged_all.rds"

  seed: 42
---

# Description

Compute QC metrics and doublet status for a collection of Seurat objects, then merge them into a single object for downstream analyses.

This template is designed to work with the outputs of the previous step (00_CreateSeuratFromCounts), which can produce:
- a list RDS file `seurat_objects_by_sample.rds`, and/or  
- per sample Seurat objects `<sample_id>.rds` plus a sample metadata table.

## RMarkdown Setup

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo = TRUE}

library(Seurat)
library(DoubletFinder)
library(sctransform)
library(scCustomize)
library(speckle)
library(scater)
library(dplyr)

```


### Parameters

Parameters used to generate this report:

``` {r print_parameters, echo=TRUE}

str(params)

```


### Functions

```{r load_functions, echo=TRUE}

# Source helper functions relative to the project root

bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c("qc.R")
)

```

## Load Seurat objects

We support two input modes:

- Use a list RDS containing all Seurat objects, as produced by 00_CreateSeuratFromCounts.

- Load individual Seurat RDS files using the sample table and seurat_objects_dir.

```{r load_seurat_objs, echo=TRUE}

# Writing
seurat_list <- list()

if (isTRUE(params$use_seurat_list)) {
  
  cat("Mode 1: load list RDS\n")
  
  if (!file.exists(params$seurat_list_rds)) {
    stop("use_seurat_list is TRUE but seurat_list_rds does not exist: ",
         params$seurat_list_rds)
  }
  seurat_list <- readRDS(params$seurat_list_rds)
  if (!is.list(seurat_list) || length(seurat_list) == 0) {
    stop("seurat_list_rds does not contain a nonempty list.")
  }
  cat("Loaded Seurat object list from:", params$seurat_list_rds, "\n")
  cat("Samples:", paste(names(seurat_list), collapse = ", "), "\n")
  
} else {
  
  cat("Mode 2: load per sample RDS using the sample table\n")
  
  sample_table <- read.table(
    file = params$sample_table,
    sep = params$sample_table_sep,
    header = TRUE,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  
  cat("Sample table loaded with", nrow(sample_table), "rows.\n")
  
  sample_id_col <- params$sample_id_col
  if (is.null(sample_id_col) || !nzchar(sample_id_col)) {
    stop("params$sample_id_col is empty or NULL.")
  }
  if (!sample_id_col %in% colnames(sample_table)) {
    stop("sample_id_col '", sample_id_col, "' not found in sample_table.")
  }
  
  sample_ids <- as.character(sample_table[[sample_id_col]])
  
  if (!dir.exists(params$seurat_objects_dir)) {
    stop("seurat_objects_dir does not exist: ", params$seurat_objects_dir)
  }
  
  for (sid in sample_ids) {
    rds_file <- file.path(params$seurat_objects_dir, paste0(sid, ".rds"))
    if (!file.exists(rds_file)) {
      stop("RDS file for sample '", sid, "' not found at: ", rds_file)
    }
    cat("Loading Seurat object for sample:", sid, "from", rds_file, "\n")
    srt <- readRDS(rds_file)
    seurat_list[[sid]] <- srt
  }
}

cat("Basic summary\n")

summary_df <- data.frame(
  sample_id = names(seurat_list),
  n_features = sapply(seurat_list, nrow),
  n_cells = sapply(seurat_list, ncol),
  stringsAsFactors = FALSE
)
summary_df

```

## Compute QC metrics

```{r compute_QC_metrics}

cat("Compute QC metrics (isOutlier) for each sample\n")

QC_params <- list(
  QC_proxy_vars = params$qc_proxy_vars,
  nmad_lower = params$nmad_lower,
  nmad_upper = params$nmad_upper,
  log = params$qc_log
)

for (sid in names(seurat_list)) {
  cat("\nRunning QCReportFilter on sample:", sid, "\n")
  seurat_list[[sid]] <- QCReportFilter(
    srt_obj = seurat_list[[sid]],
    params = QC_params
  )
}


if (isTRUE(params$run_doublet_detection)) {
  cat("Doublet detection using DoubletFinder\n")
  
  doublet_detection_params <- list(
    diet_seurat = params$doublet_diet_seurat
  )
  
  for (sid in names(seurat_list)) {
    cat("\nRunning DoubletDetection on sample:", sid, "\n")
    seurat_list[[sid]] <- DoubletDetection(
      srt_obj = seurat_list[[sid]],
      params = doublet_detection_params
    )
  }
  
} else {
  cat("run_doublet_detection is FALSE; skipping DoubletFinder.\n")
}

cat("Save per sample Seurat objects (updated)\n")

```

## Saving seurat objects

The seurat objects now have QC data that can be visualized with 03_QCReport.Rmd

```{r save_updated_seurat_objs, echo=TRUE}

# Reuse seurat_objects_dir if available; otherwise derive from merged_raw_rds directory

out_dir <- if (!is.null(params$seurat_objects_dir) && nzchar(params$seurat_objects_dir)) {
  params$seurat_objects_dir
} else {
  dirname(params$merged_raw_rds)
}

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

for (sid in names(seurat_list)) {
  out_file <- file.path(out_dir, paste0(sid, ".rds"))
  saveRDS(seurat_list[[sid]], file = out_file)
  cat("Saved updated Seurat object for sample", sid, "to", out_file, "\n")
}

if (!is.null(params$merged_raw_rds)) {

cat("Merging Seurat objects\n")

if (length(seurat_list) == 0) {
  stop("No Seurat objects in seurat_list; cannot merge.")
}

samples <- names(seurat_list)

if (length(samples) == 1) {
  srt_obj <- seurat_list[[1]]
  cat("Only one sample present; merged object will be this sample.\n")
} else {
  
#  merge() accepts a list in y and add.cell.ids
  srt_obj <- merge(
    x = seurat_list[[1]],
    y = seurat_list[-1],
    add.cell.ids = samples,
    project = "scRNA_Seq_project"
  )
}

cat("Merged object dimensions (features x cells):\n")
print(dim(srt_obj))

# Create filtered version based on isOutlier if available

cat("Saving merged objects\n")

dir.create(dirname(params$merged_raw_rds), recursive = TRUE, showWarnings = FALSE)

saveRDS(srt_obj, file = params$merged_raw_rds)

cat("Saved merged object to :", params$merged_raw_rds, "\n")

}

```

## Session info

```{r session_info, echo=TRUE}

sessionInfo()

```