---
title: "02_FilterPlotQCData"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  seurat_obj: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/merged_all.rds"
  sample_id_col: "sample_id"
  save_filtered_object: TRUE
  save_qc_figures: TRUE
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/"
  seed: 42
---

# Description

Filter and visualize QC metrics.

This report loads the merged Seurat object, visualizes QC metrics before filtering,
filters cells flagged as outliers, and generates publication-ready QC plots.

---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 5,
  dpi = 300
)

set.seed(params$seed)
```

---

# Libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(dplyr)
library(ggplot2)
library(knitr)
library(patchwork)
library(scales)
```

---

# Parameters

```{r params, echo=TRUE}
str(params)
```

---

# Load helper functions

```{r load_helper, echo=TRUE}
source(file.path(params$project_root, "R", "report_utils", "qc_plot_helpers.R"))
```

---

# Prepare output directory

```{r create_dir_structure, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
qc_fig_dir <- file.path(params$out_dir, "qc_figures")
dir.create(qc_fig_dir, recursive = TRUE, showWarnings = FALSE)
obj_dir <- file.path(params$out_dir, "objs")
dir.create(obj_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# Load Seurat object

```{r load_seurat, echo=TRUE}
srt <- readRDS(params$seurat_obj)

cat("Assays:\n")
print(Assays(srt))

cat("\nDimensions:\n")
print(dim(srt))
```

---

# QC before filtering

## Violin plots

```{r violin_plot_before_filtering, echo=TRUE}
qc_violin_before_nCount_RNA <- make_qc_violin_plot(
  srt,
  c("nCount_RNA")
)

qc_violin_before_nFeature_RNA <- make_qc_violin_plot(
  srt,
  c("nFeature_RNA")
)

qc_violin_before_percent_mito <- make_qc_violin_plot(
  srt,
  c("percent.mito")
)

qc_violin_before_nCount_RNA
qc_violin_before_nFeature_RNA
qc_violin_before_percent_mito

```

## Scatter plots

```{r scatter_plot_before_filtering, echo=TRUE}
scatter1 <- make_feature_scatter(
  srt,
  "nCount_RNA",
  "percent.mito",
  "Before filtering: counts vs mito"
)

scatter2 <- make_feature_scatter(
  srt,
  "nCount_RNA",
  "nFeature_RNA",
  "Before filtering: counts vs features"
)

scatter1
scatter2
```

---

# Apply filtering

```{r filtering, echo=TRUE}
cells_before <- ncol(srt)

srt_filtered <- subset(srt, subset = isOutlier == FALSE)

cells_after <- ncol(srt_filtered)
```

---

## Filtering summary

```{r filtering_summary, echo=TRUE}

filtering_df <- data.frame(
  step = c("Before filtering", "After filtering"),
  n_cells = c(cells_before, cells_after)
)

filtering_df$percent_retained <- round(
  100 * filtering_df$n_cells / filtering_df$n_cells[1],
  2
)

kable(filtering_df, caption = "Effect of QC filtering on cell counts")
```

---

## Filtering summary per sample

```{r per_sample_filtering, echo=TRUE}

sample_counts_before <- srt@meta.data %>%
  count(.data[[params$sample_id_col]], name = "cells_before")

sample_counts_after <- srt_filtered@meta.data %>%
  count(.data[[params$sample_id_col]], name = "cells_after")

sample_filter_df <- left_join(
  sample_counts_before,
  sample_counts_after,
  by = params$sample_id_col
)

sample_filter_df$percent_retained <- round(
  100 * sample_filter_df$cells_after / sample_filter_df$cells_before,
  2
)

kable(sample_filter_df, caption = "Per-sample cell retention after QC filtering")
```

# QC after filtering

## Violin plots

```{r violin_plot_after_filtering, echo=TRUE}

qc_violin_after_nCount_RNA <- make_qc_violin_plot(
  srt_filtered,
  c("nCount_RNA")
)

qc_violin_after_nFeature_RNA <- make_qc_violin_plot(
  srt_filtered,
  c("nFeature_RNA")
)

qc_violin_after_percent_mito <- make_qc_violin_plot(
  srt_filtered,
  c("percent.mito")
)

qc_violin_after_nCount_RNA
qc_violin_after_nFeature_RNA
qc_violin_after_percent_mito


```

## Scatter plots

```{r scatter_plot_after_filtering, echo=TRUE}
scatter_after1 <- make_feature_scatter(
  srt_filtered,
  "nCount_RNA",
  "percent.mito",
  "After filtering: counts vs mito"
)

scatter_after2 <- make_feature_scatter(
  srt_filtered,
  "nCount_RNA",
  "nFeature_RNA",
  "After filtering: counts vs features"
)

scatter_after1
scatter_after2
```

---

# Save outputs

```{r save_outputs, echo=TRUE}
if (params$save_filtered_object) {
  saveRDS(
    srt_filtered,
    file.path(obj_dir, "srt_obj_filtered.rds")
  )
}

if (params$save_qc_figures) {
  save_qc_plot(qc_violin_before_nCount_RNA, "qc_violin_before_nCount_RNA.pdf", qc_fig_dir)
  save_qc_plot(qc_violin_before_nFeature_RNA, "qc_violin_before_nFeature_RNA.pdf", qc_fig_dir)
  save_qc_plot(qc_violin_before_percent_mito, "qc_violin_before_percent_mito.pdf", qc_fig_dir)
  save_qc_plot(qc_violin_after_nCount_RNA, "qc_violin_after_nCount_RNA.pdf", qc_fig_dir)
  save_qc_plot(qc_violin_after_nFeature_RNA, "qc_violin_after_nFeature_RNA.pdf", qc_fig_dir)
  save_qc_plot(qc_violin_after_percent_mito, "qc_violin_after_percent_mito.pdf", qc_fig_dir)
  save_qc_plot(scatter1 + scatter2, "qc_scatter_before.pdf", qc_fig_dir)
  save_qc_plot(scatter_after1 + scatter_after2, "qc_scatter_after.pdf", qc_fig_dir)
}
```

---

# Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
