---
title: "02_FilterPlotQCData"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  seurat_obj: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/merged_all.rds"
  sample_id_col: "sample_id"
  save_filtered_object: TRUE
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"
  seed: 42
---

# Description

Filter and visualize QC metrics.

This report is designed to work with the outputs of **01_ComputeQCMetrics.Rmd** and performs the following tasks:

- Load the merged Seurat object containing QC annotations
- Visualize QC metrics **before filtering**
- Filter cells flagged as outliers (`isOutlier == TRUE`)
- Produce QC summaries and plots **after filtering**
- Save a filtered Seurat object

The goal of this report is purely **QC visualization and filtering**, keeping it logically separate from normalization or SCTransform.

---

# RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)
```

---

# Preparing the environment

## Libraries

```{r load_libraries}
library(Seurat)
library(dplyr)
library(ggplot2)
library(knitr)
```

---

# Parameters

```{r print_parameters}
str(params)
```

---

# Load helper functions

```{r load_functions}

bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R"
  )
)
```

---

# Prepare output directory

```{r prepare_output_dir}

dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)

cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n")
```

---

# Load Seurat object

```{r load_input_object}

input_srt <- read_seurat_rds(params$seurat_obj)

cat("Input object loaded from:\n")
cat(params$seurat_obj, "\n\n")

cat("Assays present:\n")
print(Assays(input_srt))

cat("\nDimensions (features x cells):\n")
print(dim(input_srt))
```

---

# Metadata validation

```{r metadata_validation}

required_cols <- c(
  params$sample_id_col,
  "isOutlier",
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mito"
)

missing_cols <- setdiff(required_cols, colnames(input_srt@meta.data))

if (length(missing_cols) > 0) {
  stop("Missing required metadata columns: ", paste(missing_cols, collapse = ", "))
}
```

---

# Input object summary

```{r input_summary}

summary_df <- data.frame(
  metric = c("n_features", "n_cells", "n_samples"),
  value = c(
    nrow(input_srt),
    ncol(input_srt),
    length(unique(input_srt@meta.data[[params$sample_id_col]]))
  )
)

kable(summary_df, caption = "Input Seurat object summary")
```

---

# QC metrics before filtering

```{r qc_plots_before, fig.width=12}

VlnPlot(
  input_srt,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mito"),
  ncol = 3,
  pt.size = 0
)
```

```{r qc_scatter_before}

FeatureScatter(input_srt, feature1 = "nCount_RNA", feature2 = "percent.mito")
FeatureScatter(input_srt, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

---

# Apply QC filtering

```{r filter_cells}

cells_before <- ncol(input_srt)

srt_obj <- subset(
  input_srt,
  subset = isOutlier == FALSE
)

cells_after <- ncol(srt_obj)
```

---

# Filtering summary

```{r filtering_summary}

filtering_df <- data.frame(
  step = c("Before filtering", "After filtering"),
  n_cells = c(cells_before, cells_after)
)

filtering_df$percent_retained <- round(
  100 * filtering_df$n_cells / filtering_df$n_cells[1],
  2
)

kable(filtering_df, caption = "Effect of QC filtering on cell counts")
```

---

# Filtering summary per sample

```{r per_sample_filtering}

sample_counts_before <- input_srt@meta.data %>%
  count(.data[[params$sample_id_col]], name = "cells_before")

sample_counts_after <- srt_obj@meta.data %>%
  count(.data[[params$sample_id_col]], name = "cells_after")

sample_filter_df <- left_join(
  sample_counts_before,
  sample_counts_after,
  by = params$sample_id_col
)

sample_filter_df$percent_retained <- round(
  100 * sample_filter_df$cells_after / sample_filter_df$cells_before,
  2
)

kable(sample_filter_df, caption = "Per-sample cell retention after QC filtering")
```

---

# QC metrics after filtering

```{r qc_plots_after, fig.width=12}

VlnPlot(
  srt_obj,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mito"),
  ncol = 3,
  pt.size = 0
)
```

```{r qc_scatter_after}

FeatureScatter(srt_obj, feature1 = "nCount_RNA", feature2 = "percent.mito")
FeatureScatter(srt_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

---

# Save filtered Seurat object

```{r save_filtered_object}

filtered_file <- file.path(params$out_dir, "srt_obj_filtered.rds")

if (isTRUE(params$save_filtered_object)) {
  saveRDS(srt_obj, file = filtered_file)
  cat("Filtered Seurat object saved to:\n", filtered_file, "\n")
}
```

---

# Final object summary

```{r final_summary}

final_df <- data.frame(
  metric = c("n_features", "n_cells"),
  value = c(
    nrow(srt_obj),
    ncol(srt_obj)
  )
)

kable(final_df, caption = "Filtered Seurat object summary")
```

---

# Session info

```{r session_info}
sessionInfo()
```
