---
title: "03_ComputeSCT"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  seurat_obj: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/merged_all.rds"
  sample_id_col: "sample_id"
  save_intermediate_files: TRUE
  
  # Output directory
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"

  seed: 42
---

# Description

Run SCTransform.

This template is designed to work with the outputs of the previous step (02_FilterPlotQCData.Rmd) and will produce:

- a Seurat object with normalized and scaled RNA assay

- a list of per-sample Seurat objects with SCT assay

- a merged Seurat object containing RNA and SCT assays

## RMarkdown Setup

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)

```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo = TRUE}

library(Seurat)
library(sctransform)
library(dplyr)
library(ggplot2)
library(knitr)

```


### Parameters

Parameters used to generate this report:

``` {r print_parameters, echo=TRUE}

str(params)

```
## Environment configuration

```{r load_functions, echo=TRUE}

# Source helper functions relative to the project root

bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R",
    "integration_sct.R"
  )
)

```

### Create output directory 

```{r prepare_output_dir, echo=TRUE}

dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n")

```

### Load Seurat object

```{r load_input_object, echo=TRUE}

input_srt <- read_seurat_rds(params$seurat_obj)

cat("Input object loaded from:\n")
cat(params$seurat_obj, "\n\n")

cat("Assays present:\n")
print(Assays(input_srt))

cat("\nDimensions (features x cells):\n")
print(dim(input_srt))

```

### Seurat object structure

```{r input_metadata_summary, echo=TRUE}
required_cols <- c(params$sample_id_col, "isOutlier", "nCount_RNA", "nFeature_RNA", "percent.mito")
missing_cols <- setdiff(required_cols, colnames(input_srt@meta.data))

if (length(missing_cols) > 0) {
  stop("Missing required metadata columns: ", paste(missing_cols, collapse = ", "))
}

input_summary <- data.frame(
  metric = c("n_features", "n_cells", "n_samples"),
  value = c(
    nrow(input_srt),
    ncol(input_srt),
    length(unique(input_srt@meta.data[[params$sample_id_col]]))
  )
)

kable(input_summary, caption = "Basic summary of the input Seurat object")

```

## Normalize and scale RNA assay

```{r normalize_scale_RNA_assay, echo=TRUE}

scaled_rds_file <- file.path(params$out_dir, "srt_obj_scaled.rds")

if (file.exists(scaled_rds_file)) {
  srt_obj <- readRDS(file = scaled_rds_file)
  cat("Reusing previously saved scaled object:\n")
  cat(scaled_rds_file, "\n")

} else {

  srt_obj <- input_srt

  if (!"isOutlier" %in% colnames(srt_obj@meta.data)) {
    stop("Column 'isOutlier' not found in Seurat object metadata.")
  }

  n_before <- ncol(srt_obj)

  # Remove outliers
  srt_obj <- subset(x = srt_obj, subset = isOutlier == FALSE)

  n_after <- ncol(srt_obj)

  # Join RNA layers if needed
  srt_obj <- JoinLayers(srt_obj, assay = "RNA")

  # Normalize RNA assay
  srt_obj <- NormalizeData(
    object = srt_obj,
    assay = "RNA",
    normalization.method = "LogNormalize",
    scale.factor = 10000
  )

  # Scale RNA assay
  srt_obj <- ScaleData(
    object = srt_obj,
    assay = "RNA",
    vars.to.regress = "percent.mito"
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(srt_obj, file = scaled_rds_file)
  }
}

```

### Filtering summary

```{r filtering_summary, echo=TRUE}
filtering_df <- data.frame(
  step = c("Before outlier filtering", "After outlier filtering"),
  n_cells = c(ncol(input_srt), ncol(srt_obj))
)

filtering_df$percent_retained <- round(100 * filtering_df$n_cells / filtering_df$n_cells[1], 2)

kable(filtering_df, caption = "Effect of outlier filtering on cell count")

```

### Filtering summary per sample

```{r filtered_per_sample_table, echo=TRUE}
filtered_sample_df <- srt_obj@meta.data %>%
  count(.data[[params$sample_id_col]], name = "n_cells") %>%
  arrange(desc(n_cells))

kable(filtered_sample_df, caption = "Number of retained cells per sample after outlier filtering")

```

```{r filtered_qc_plots, echo=TRUE, fig.height=4, fig.width=12}
VlnPlot(
  srt_obj,
  features = c("nCount_RNA", "nFeature_RNA", "percent.mito"),
  ncol = 3,
  pt.size = 0
)

```

```{r compute_SCT, echo=TRUE}

sct_list_file <- file.path(params$out_dir, "sct_list.rds")
merged_sct_file <- file.path(params$out_dir, "srt_obj_norm_scaled_RNA_plus_SCT.rds")

# Reuse SCT outputs if available
if (file.exists(sct_list_file) && file.exists(merged_sct_file)) {

  sample_list <- readRDS(sct_list_file)
  srt_obj <- readRDS(merged_sct_file)

} else {

  # Split by sample and run SCT per sample
  sample_list <- split_seurat(srt_obj, split_by = params$sample_id_col)

  sample_list <- sct_transform_list(
    obj_list = sample_list,
    vst.flavor = "v2",
    variable.features.n = 10000,
    method = "glmGamPoi",
    vars_to_regress = "percent.mito",
    return_only_var_genes = FALSE,
    verbose = TRUE
  )

  srt_obj <- merge(
    x = sample_list[[1]],
    y = sample_list[-1],
    add.cell.ids = names(sample_list),
    project = "SCTransform"
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(sample_list, file = sct_list_file)
    saveRDS(srt_obj, file = merged_sct_file)
  }
}

gc()

```


```{r session_info, echo=TRUE}

sessionInfo()

```
