---
title: "03_ComputeSCT"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  seurat_obj: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_filtered.rds"
  sample_id_col: "sample_id"
  save_intermediate_files: TRUE

  # Output directory
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"

  seed: 42
---

# Description

Run SCTransform.

This template is designed to work with the outputs of the previous step (**02_FilterPlotQCData.Rmd**) and will produce:

- a Seurat object with normalized and scaled RNA assay
- a list of per-sample Seurat objects with SCT assay
- a merged Seurat object containing RNA and SCT assays

In addition to computing the objects, this report also summarizes the input data and provides diagnostic tables and figures for the RNA normalization/scaling and SCT steps.

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(sctransform)
library(dplyr)
library(ggplot2)
library(knitr)
library(patchwork)
```

### Parameters

Parameters used to generate this report:

```{r print_parameters, echo=TRUE}
str(params)
```

### Load helper functions

```{r load_functions, echo=TRUE}
# Source helper functions relative to the project root

bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R",
    "integration_sct.R"
  )
)
```

### Create output directory 

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n")
```

## Load Seurat object

```{r load_input_object, echo=TRUE}
input_srt <- read_seurat_rds(params$seurat_obj)

cat("Input object loaded from:\n")
cat(params$seurat_obj, "\n\n")

cat("Assays present:\n")
print(Assays(input_srt))

cat("\nDimensions (features x cells):\n")
print(dim(input_srt))
```

### Input object summary

```{r input_metadata_summary, echo=TRUE}
required_cols <- c(params$sample_id_col, "nCount_RNA", "nFeature_RNA", "percent.mito")
missing_cols <- setdiff(required_cols, colnames(input_srt@meta.data))

if (length(missing_cols) > 0) {
  stop("Missing required metadata columns: ", paste(missing_cols, collapse = ", "))
}

input_summary <- data.frame(
  metric = c("n_features", "n_cells", "n_samples"),
  value = c(
    nrow(input_srt),
    ncol(input_srt),
    length(unique(input_srt@meta.data[[params$sample_id_col]]))
  )
)

kable(input_summary, caption = "Basic summary of the input Seurat object")
```

```{r per_sample_input_summary, echo=TRUE}
sample_summary <- input_srt@meta.data %>%
  count(.data[[params$sample_id_col]], name = "n_cells") %>%
  arrange(desc(n_cells))

kable(sample_summary, caption = "Number of cells per sample in the input Seurat object")
```

## Normalize and scale RNA assay

```{r normalize_scale_RNA_assay, echo=TRUE}
scaled_rds_file <- file.path(params$out_dir, "srt_obj_scaled.rds")

if (file.exists(scaled_rds_file)) {
  srt_obj <- readRDS(file = scaled_rds_file)
  cat("Reusing previously saved scaled object:\n")
  cat(scaled_rds_file, "\n")
} else {
  srt_obj <- input_srt

  # Join RNA layers if needed
  srt_obj <- JoinLayers(srt_obj, assay = "RNA")

  # Normalize RNA assay
  srt_obj <- NormalizeData(
    object = srt_obj,
    assay = "RNA",
    normalization.method = "LogNormalize",
    scale.factor = 10000
  )

  # Scale RNA assay
  srt_obj <- ScaleData(
    object = srt_obj,
    assay = "RNA",
    block.size = 5000,
    vars.to.regress = "percent.mito"
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(srt_obj, file = scaled_rds_file)
  }
}
```

### RNA assay summary after normalization and scaling

```{r normalized_summary, echo=TRUE}
normalized_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay"),
  value = c(
    nrow(srt_obj),
    ncol(srt_obj),
    DefaultAssay(srt_obj)
  )
)

kable(normalized_summary, caption = "Summary of the Seurat object after RNA normalization and scaling")
```

## Compute SCT per sample

```{r split_object_summary, echo=TRUE}
sample_list <- split_seurat(srt_obj, split_by = params$sample_id_col)

split_summary <- data.frame(
  sample_id = names(sample_list),
  n_features = vapply(sample_list, nrow, numeric(1)),
  n_cells = vapply(sample_list, ncol, numeric(1)),
  stringsAsFactors = FALSE
) %>%
  arrange(desc(n_cells))

kable(split_summary, caption = "Per-sample Seurat objects before SCTransform")
```

```{r compute_SCT, echo=TRUE}
sct_list_file <- file.path(params$out_dir, "sct_list.rds")
merged_sct_file <- file.path(params$out_dir, "srt_obj_norm_scaled_RNA_plus_SCT.rds")

# Reuse SCT outputs if available
if (file.exists(sct_list_file) && file.exists(merged_sct_file)) {
  sample_list <- readRDS(sct_list_file)
  srt_obj <- readRDS(merged_sct_file)
  cat("Reusing previously saved SCT outputs:\n")
  cat(sct_list_file, "\n")
  cat(merged_sct_file, "\n")
} else {
  sample_list <- sct_transform_list(
    obj_list = sample_list,
    vst.flavor = "v2",
    variable.features.n = 10000,
    method = "glmGamPoi",
    vars_to_regress = "percent.mito",
    return_only_var_genes = FALSE,
    verbose = TRUE
  )

  srt_obj <- merge(
    x = sample_list[[1]],
    y = sample_list[-1],
    add.cell.ids = names(sample_list),
    project = "SCTransform"
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(sample_list, file = sct_list_file)
    saveRDS(srt_obj, file = merged_sct_file)
  }
}

gc()
```

### SCT output summaries

```{r sct_output_summary, echo=TRUE}
sct_summary <- data.frame(
  sample_id = names(sample_list),
  n_features = vapply(sample_list, nrow, numeric(1)),
  n_cells = vapply(sample_list, ncol, numeric(1)),
  assays = vapply(sample_list, function(x) paste(Assays(x), collapse = ", "), character(1)),
  stringsAsFactors = FALSE
) %>%
  arrange(desc(n_cells))

kable(sct_summary, caption = "Per-sample Seurat objects after SCTransform")
```

```{r merged_sct_summary, echo=TRUE}
merged_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay", "assays"),
  value = c(
    nrow(srt_obj),
    ncol(srt_obj),
    DefaultAssay(srt_obj),
    paste(Assays(srt_obj), collapse = ", ")
  )
)

kable(merged_summary, caption = "Summary of the merged Seurat object after SCTransform")
```

### Diagnostic figures for SCT results

```{r sct_violin_merged, echo=TRUE, fig.width=12, fig.height=4.2}
DefaultAssay(srt_obj) <- "SCT"

VlnPlot(
  srt_obj,
  features = c("nCount_SCT", "nFeature_SCT"),
  ncol = 2,
  pt.size = 0
)
```

```{r variable_features_example, echo=TRUE, fig.width=10, fig.height=4.5}
example_sample <- sample_list[[1]]
DefaultAssay(example_sample) <- "SCT"

var_plot <- VariableFeaturePlot(example_sample)
var_plot + ggtitle(paste0("Variable features for sample: ", names(sample_list)[1]))
```

```{r pca_example_sample, echo=TRUE, eval=FALSE}
# Optional diagnostic: run only if you want a PCA plot for one SCT-transformed sample.
# example_sample <- RunPCA(example_sample, assay = "SCT", verbose = FALSE)
DimPlot(example_sample, reduction = "pca") + ggtitle(paste0("PCA of sample: ", names(sample_list)[1]))
```

## Output files

```{r output_files_summary, echo=TRUE}
output_files <- data.frame(
  object = c(
    "Scaled RNA Seurat object",
    "List of SCT-transformed Seurat objects",
    "Merged RNA + SCT Seurat object"
  ),
  file = c(
    scaled_rds_file,
    sct_list_file,
    merged_sct_file
  ),
  exists = file.exists(c(scaled_rds_file, sct_list_file, merged_sct_file)),
  stringsAsFactors = FALSE
)

kable(output_files, caption = "Expected output files produced by this report")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
