---
title: "02_IntegrateSeurat"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  seurat_obj: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/merged_all.rds"
  sample_id_col: "sample_id"
  save_intermediate_files: TRUE # save intermediate files
  
  # path to table with cell cycle genes
  cell_cycle_gene_table: "../data/metadata/cell_cycle_genes.tsv"

  # Output directory
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"

  seed: 42
---

# Description

Integrate


This template is designed to work with the outputs of the previous step (00_CreateSeuratFromCounts), which can produce:
- a list RDS file `seurat_objects_by_sample.rds`, and/or  
- per sample Seurat objects `<sample_id>.rds` plus a sample metadata table.

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)

```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo = TRUE}

library(Seurat)
library(sctransform)
library(scCustomize)
library(DT)
library(stringr)
library(ggplot2)

#library(parallel)
#library(future)
#library(future.apply)

```


### Parameters

Parameters used to generate this report:

``` {r print_parameters, echo=TRUE}

params <- list(
  project_root = "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas",
  seurat_obj = "...",
  sample_id_col = "sample_ID",
  cell_cycle_gene_table = "...",
  seurat_out_obj = "...",
  seed = 42,
  out_dir = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"
)

str(params)

```

## Environment configuration

```{r load_functions, echo=TRUE}

# Source helper functions relative to the project root

bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R"
  )
)

```

```{r process_seurat_obj, echo=TRUE}

# Load merged object
#srt_obj <- read_seurat_rds(params$seurat_obj)

# Remove outliers
#srt_obj <- subset(x = srt_obj, isOutlier == FALSE)

# Join layers
#srt_obj <- JoinLayers(srt_obj)

# Normalize
#srt_obj <- NormalizeData(srt_obj,
#                         assay = "RNA",
#                         normalization.method = "LogNormalize",
#                         scale.factor = 10000)
# Scale
#srt_obj <- ScaleData(srt_obj,
#                     assay = "RNA",
#                     vars.to.regress = "percent.mito"
#)

if (params$save_intermediate_files == TRUE) {
#  saveRDS(srt_obj, file = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_scaled_tmp.rds")
}

#srt_obj <- readRDS(file = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_scaled_tmp.rds")


# Load cell cycle table and derive gene sets
#cell_cycle_df <- read_cell_cycle_table(params$cell_cycle_gene_table)
#cc_feats <- get_cell_cycle_features(cell_cycle_df)
#
# Score cell cycle
#srt_obj <- CellCycleScoring(object = srt_obj,
#                            s.features = cc_feats$s_features,
#                            g2m.features = cc_feats$g2m_features)

# Split by sample and run SCT per sample

#sample_list <- split_seurat(srt_obj, split_by = params$sample_id_col)

#sample_list <- lapply(X = sample_list, 
#                      FUN = SCTransform,
#                      vst.flavor = "v2",
#                      variable.features.n = 10000,
#                      method = "glmGamPoi",
#                      vars.to.regress = c("percent.mito"),
#                      return.only.var.genes = FALSE)

# sample_list <- sct_transform_list(
#   sample_list,
#   method = "glmGamPoi",
# #  vars_to_regress = c("percent.mito", "S.Score", "G2M.Score"),
#   vars_to_regress = c("percent.mito"),
#   return_only_var_genes = FALSE
# )

if (params$save_intermediate_files == TRUE) {
#  saveRDS(sample_list, file = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/sct_list_v2_tmp.rds")
}

sample_list <- readRDS(file = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/sct_list_v2_tmp.rds")

#srt_obj <- merge(x = sample_list[[1]], y = sample_list[-1])

if (params$save_intermediate_files == TRUE) {
#  saveRDS(srt_obj, file = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_norm_scaled_RNA_plus_SCT_v2.rds")
}

#srt_obj <- readRDS(file = "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_norm_scaled_RNA_plus_SCT_v2.rds")

#gc()

```

## Integration by stage

Now we integrate each stage individually to see how the major cell types (based on module scores) will group

```{r integrate_stages, echo = TRUE}

# Integration features
features <- SelectIntegrationFeatures(object.list = sample_list, nfeatures = 3000)

# PCA
sample_list <- lapply(
  X = sample_list,
  FUN = function(x) {
    x <- RunPCA(x, verbose = FALSE, features = features)
    return(x)
  }
)

gc()

# integrate by stage
stages <- c("0-1_days", "3_days", "7_days", "Adult")

stage_obj_list <- list()

# integrate each stage
for (stage_id in stages) {
  print(paste0("Integrating stage ", stage_id))
  
  # Subset by stage
  filtered_list <- Filter(
    function(obj) {
      # Extract the metadata column
      stage_values <- obj@meta.data$Stage
      
      # Since all cells share the same value, check the first one
      stage_values[1] == stage_id
    },
    sample_list
  )
  
#  features <- SelectIntegrationFeatures(object.list = filtered_list, nfeatures = 3000)
  
  # filtered_list <- lapply(
  #   X = filtered_list,
  #   FUN = function(x) {
  #     #    x <- SCTransform(x, verbose = FALSE)   # avoid regressing cell cycle here unless you are sure
  #     x <- RunPCA(x, verbose = FALSE, features = features)
  #     return(x)
  #   }
  # )
  
  options(future.globals.maxSize = 300 * 1024^3) # Sets the limit to 300 GiB
  filtered_list <- PrepSCTIntegration(object.list = filtered_list, anchor.features = features, verbose = FALSE)

  gc()

  anchors <- FindIntegrationAnchors(
    object.list = filtered_list,
    normalization.method = "SCT",
    anchor.features = features,
    reduction = "rpca",
    dims = 1:40
  )

  gc()

# ----------------------------
# 3) Integrate, then continue standard downstream
# ----------------------------
  atlas_int <- IntegrateData(
    anchorset = anchors,
    normalization.method = "SCT",
    dims = 1:40
  )

  gc()

  DefaultAssay(atlas_int) <- "integrated"
  atlas_int <- RunPCA(atlas_int, npcs = 40, verbose = FALSE)
  atlas_int <- RunUMAP(atlas_int, reduction = "pca", dims = 1:40)
  atlas_int <- FindNeighbors(atlas_int, reduction = "pca", dims = 1:40)
  atlas_int <- FindClusters(atlas_int, resolution = c(0.6, 1.2, 1.8, 2.4, 4), algorithm = 4)
  
  stage_name <- str_replace(stage_id, pattern = "-", replacement = "_")
  
  stage_obj_list[stage_name] <- atlas_int
  
  out_file_name <- paste0(params$out_dir, "srt_", stage_name, "_integrated.rds")
  print(paste0("Saving file ", out_file_name))  
  saveRDS(atlas_int, file = out_file_name)  
  print("Done")
  gc()
}

out_file_name <- paste0(params$out_dir, "srt_list_integrated.rds")
saveRDS(stage_obj_list, file = out_file_name)  

out_file_name <- paste0(params$out_dir, "srt_list_integration_features.rds")
saveRDS(features, file = out_file_name)  

```
## Add module scores from the literature
This step adds module scores for major cell types. All were gathered from the literature and have been validated from M. lignano, with the exception of epidermal markers, which were selected from S. polychroa and mapped to M. lignano homologs by blast (https://doi.org/10.1126/science.aaq1736).

### Import table of marker genes
```{r import_module_score_table, echo = TRUE}

marker_genes <- read.table("~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas/data/metadata/literature_marker_genes.tsv", sep = "\t", header = TRUE, quote = "")

# Seurat doesn't allow underscores in gene symbols and has replaced them in the object, so we need to replace them here as well
marker_genes$Gene_ID <- str_replace(marker_genes$Gene_ID, pattern = "_", replacement = "-")

datatable(marker_genes)

```

### Add module scores
```{r add_module_scores, echo = TRUE}

#in_file_name <- paste0(params$out_dir, "srt_list_integrated.rds")

#stage_obj_list <- readRDS(file = in_file_name)  

anchor_cell_of_adhesive_organ <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "anchor_cell_of_adhesive_organ"]
cement_gland_cell <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "cement_gland_cell"]
epidermal_secretory_cell <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "epidermal_secretory_cell"]
female_antrum <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "female_antrum"]
female_germline_ovary <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "female_germline_ovary"]
gut <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "gut"]
gut_specialized_cell <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "gut_specialized_cell"]
male_germline_testis <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "male_germline_testis"]
muscle <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "muscle"]
nervous_system <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "nervous_system"]
prostate <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "prostate"]
rhabdite <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "rhabdite-containing_cell"]
secretory_cell_of_adhesive_organ <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "secretory_cell_of_adhesive_organ"]
stem_cells_progenitors <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "stem_cells_progenitors"]
epidermal <- marker_genes$Gene_ID[marker_genes$Cell_tissue_type == "epidermal"]

marker_genes_list <- list(anchor_cell_of_adhesive_organ,
                          cement_gland_cell,
                          epidermal_secretory_cell,
                          female_antrum,
                          female_germline_ovary,
                          gut,
                          gut_specialized_cell,
                          male_germline_testis,
                          muscle,
                          nervous_system,
                          prostate,
                          rhabdite,
                          secretory_cell_of_adhesive_organ,
                          stem_cells_progenitors,
                          epidermal)

names(marker_genes_list) <- c("anchor_cell_of_adhesive_organ",
                              "cement_gland_cell",
                              "epidermal_secretory",
                              "female_antrum",
                              "female_germline_ovary",
                              "gut",
                              "gut_specialized",
                              "male_germline_testis",
                              "muscle",
                              "nervous_system",
                              "prostate",
                              "rhabdite",
                              "secretory_cell_of_adhesive_organ",
                              "progenitor",
                              "epidermal")
#gc()

#stage_obj_list <- lapply(stage_obj_list, function(obj) {
#  obj <- JoinLayers(obj, assay = "RNA")
#  DefaultAssay(obj) <- "RNA"
#  obj
#})

gc()

#stage_obj_list <- lapply(stage_obj_list, function(obj) {
#  obj <- add_module_scores(obj, gene_lists = marker_genes_list)
#  obj
#})

#gc()

#out_file_name <- paste0(params$out_dir, "srt_list_integrated.rds")
#saveRDS(stage_obj_list, file = out_file_name)  

gc()

```

### Visualize module scores by stage

```{r plot_module_scores, echo = TRUE}

## 0-1 days
p1 <- FeaturePlot_scCustom(stage_obj_list[["0_1_days"]], features = names(marker_genes_list)) + ggtitle("0-1 days")

## 3 days
p2 <- FeaturePlot_scCustom(stage_obj_list[["3_days"]], features = names(marker_genes_list)) + ggtitle("3 days")

## 7 days
p3 <- FeaturePlot_scCustom(stage_obj_list[["7_days"]], features = names(marker_genes_list)) + ggtitle("7 days")

## Adult
p4 <- FeaturePlot_scCustom(stage_obj_list[["Adult"]], features = names(marker_genes_list)) + ggtitle("Adult")

print(p1)
print(p2)
print(p3)
print(p4)

```

### Integrate all samples using adult as anchor

```{r anchor_integration, echo=TRUE}

# Indices of the adult replicates in obj_list (these will be the reference)
adult_idx <- which(names(sample_list) %in% c("2_Adult_Unsorted", "2_Adult_Unsorted_DASH", "7_Sorted_Adult"))

# ----------------------------
# 1) Preprocess (SCT workflow, recommended for integration)
# ----------------------------
sample_list <- lapply(
  X = sample_list,
  FUN = function(x) {
#    x <- SCTransform(x, verbose = FALSE)   # avoid regressing cell cycle here unless you are sure
    x <- RunPCA(x, verbose = FALSE)
    return(x)
  }
)

# ----------------------------
# 2) Find anchors using adults as the reference
#    Use RPCA to be more conservative (often helps avoid over-merging)
# ----------------------------
features <- SelectIntegrationFeatures(object.list = sample_list, nfeatures = 3000)

options(future.globals.maxSize = 300 * 1024^3) # Sets the limit to 300 GiB

sample_list <- PrepSCTIntegration(object.list = sample_list, anchor.features = features, verbose = FALSE)

gc()

anchors <- FindIntegrationAnchors(
  object.list = sample_list,
  normalization.method = "SCT",
  anchor.features = features,
  reference = adult_idx,     # key line: non-adults are anchored onto adults
  reduction = "rpca",        # conservative anchors
  dims = 1:40
)

gc()

# ----------------------------
# 3) Integrate, then continue standard downstream
# ----------------------------
atlas_int <- IntegrateData(
  anchorset = anchors,
  normalization.method = "SCT",
  dims = 1:40
)

gc()

DefaultAssay(atlas_int) <- "integrated"
atlas_int <- RunPCA(atlas_int, npcs = 40, verbose = FALSE)
atlas_int <- RunUMAP(atlas_int, reduction = "pca", dims = 1:40)
atlas_int <- FindNeighbors(atlas_int, reduction = "pca", dims = 1:40)
atlas_int <- FindClusters(atlas_int, resolution = c(0.6, 1.2, 1.8, 2.4, 4), algorithm = 4)

out_file_name <- paste0(params$out_dir, "srt_integrated_all.rds")
saveRDS(atlas_int, file = out_file_name)  

gc()

```

```{r visualize_integration, echo=TRUE}
# Switch back to RNA:
DefaultAssay(atlas_int) <- "RNA"
atlas_int <- JoinLayers(atlas_int, assay = "RNA")
atlas_int <- add_module_scores(atlas_int, gene_lists = marker_genes_list)
p_module_integrated <- FeaturePlot_scCustom(atlas_int, features = names(marker_genes_list))
p_by_stage <- DimPlot_scCustom(atlas_int, group.by = "Stage")
p_split_by_stage_group_by_sample <- DimPlot_scCustom(atlas_int, split.by = "Stage", group.by = "sample_id")

p_module_integrated_tmp <- FeaturePlot_scCustom(atlas_int, features = c("male_germline_testis", "female_germline_ovary"))

saveRDS(atlas_int, file = out_file_name)

```

```{}

DefaultAssay(atlas_int) <- "RNA"

gc()

rm(obj_list)

gc()

atlas_int <- JoinLayers(atlas_int)

cols_to_keep <- c("orig.ident",
                  "nCount_RNA",
                  "nFeature_RNA",
                  "sample_ID",
                  "sample_type",
                  "percent.mito",
                  "nCount_SCT",
                  "nFeature_SCT",
                  "S.Score",
                  "G2M.Score",
                  "Phase",
                  "nCount_integrated",
                  "nFeature_integrated",
                  "integrated_snn_res.0.2",
                  "integrated_snn_res.0.4",
                  "integrated_snn_res.0.6",
                  "integrated_snn_res.0.8",
                  "integrated_snn_res.1",
                  "integrated_snn_res.1.2",
                  "integrated_snn_res.1.4",
                  "integrated_snn_res.1.6",
                  "integrated_snn_res.1.8",
                  "integrated_snn_res.2",
                  "integrated_snn_res.3",
                  "integrated_snn_res.4")

seurat_integrated_final@meta.data <- seurat_integrated_final@meta.data[, cols_to_keep, drop = FALSE]

Assays(seurat_integrated_final)

DefaultAssay(seurat_integrated_final) <- "RNA"

# Save final object
save_rds(seurat_integrated, params$seurat_out_obj)
cat("Saved integrated object to:", params$seurat_out_obj, "\n")


```


```{merge_srt_obj_assays}

# 2. Merge them in one go
seurat_integrated <- merge(
  x            = cleaned_timepoints[[1]],
  y            = cleaned_timepoints[-1],
  add.cell.ids = names(cleaned_timepoints),
  project      = "ReplicateIntegration"
)

# 3. Re‐run scaling, PCA & UMAP on the merged “integrated” assay

DefaultAssay(seurat_integrated) <- "integrated"
seurat_integrated <- FindVariableFeatures(seurat_integrated)
seurat_integrated <- ScaleData(seurat_integrated, verbose = FALSE)
seurat_integrated <- RunPCA(seurat_integrated, verbose = FALSE)
seurat_integrated <- RunUMAP(seurat_integrated, dims = 1:30)

# 5. Merge all time‐point–integrated objects in one shot
seurat_integrated <- merge(
  x            = integrated_timepoints[[1]],
  y            = integrated_timepoints[-1],
  add.cell.ids = names(integrated_timepoints),
  project      = "Timepoint_Replicate_Integration"
)



gc()


```

```{r process_integrated_srt_obj, echo=TRUE}
seurat_integrated <- FindNeighbors(seurat_integrated,
                                   reduction = "pca",
                                   dims = 1:40,
                                   verbose = FALSE)

seurat_integrated <- FindClusters(seurat_integrated,
                                  resolution = c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6, 1.8, 2, 3, 4),
                                  algorithm = 4,
                                  random.seed = 42)

saveRDS(seurat_integrated_final, file = "objs/seurat_integrated_assay_DASH.rds")

```

```{r session_info, echo=TRUE}
session

```
