---
title: "04_IntegrateSeuratObjects"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  sct_list_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/sct_list.rds"
  merged_sct_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_norm_scaled_RNA_plus_SCT.rds"
  sample_id_col: "sample_id"
  stage_col: "Stage"
  stages: !r c("0-1_days", "3_days", "7_days", "Adult")
  adult_reference_samples: !r c("2_Adult_Unsorted", "2_Adult_Unsorted_DASH", "7_Sorted_Adult")
  nfeatures: 3000
  dims_use: 40
  clustering_resolutions: !r c(0.6, 1.2, 1.8, 2.4, 4)
  save_intermediate_files: TRUE
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"
  seed: 42
---

# Description

Integrate Seurat objects.

This template is designed to work with the outputs of the previous step (**03_ComputeSCT.Rmd**) and performs two integration procedures:

- integrate replicates within each stage independently
- integrate all samples using adult samples as references

The report also provides diagnostic tables and figures to summarize inputs and integration outputs. Whenever intermediate files already exist, the report reuses them instead of recomputing from scratch.

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(sctransform)
library(scCustomize)
library(DT)
library(stringr)
library(ggplot2)
library(dplyr)
library(knitr)
library(patchwork)
```

### Parameters

Parameters used to generate this report:

```{r print_parameters, echo=TRUE}
str(params)
```

### Load helper functions

```{r load_functions, echo=TRUE}
bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R"
  )
)
```

### Prepare output directory

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n")
```

## Load data

```{r load_data, echo=TRUE}
if (!file.exists(params$sct_list_file)) {
  stop("sct_list_file not found: ", params$sct_list_file)
}
if (!file.exists(params$merged_sct_file)) {
  stop("merged_sct_file not found: ", params$merged_sct_file)
}

sample_list <- readRDS(file = params$sct_list_file)
srt_obj <- readRDS(file = params$merged_sct_file)

gc()

cat("Loaded sample_list with", length(sample_list), "objects.\n")
cat("Merged object dimensions (features x cells):\n")
print(dim(srt_obj))
```

## Input summaries

```{r input_object_summary, echo=TRUE}
input_summary <- data.frame(
  metric = c("n_samples", "merged_n_features", "merged_n_cells", "merged_assays"),
  value = c(
    length(sample_list),
    nrow(srt_obj),
    ncol(srt_obj),
    paste(Assays(srt_obj), collapse = ", ")
  ),
  stringsAsFactors = FALSE
)

kable(input_summary, caption = "Summary of the SCT inputs used for integration")
```

```{r input_sample_table, echo=TRUE}
sample_summary <- data.frame(
  sample_id = names(sample_list),
  n_features = vapply(sample_list, nrow, numeric(1)),
  n_cells = vapply(sample_list, ncol, numeric(1)),
  stage = vapply(
    sample_list,
    function(x) {
      md <- x@meta.data
      if (params$stage_col %in% colnames(md)) as.character(md[[params$stage_col]][1]) else NA_character_
    },
    character(1)
  ),
  assays = vapply(sample_list, function(x) paste(Assays(x), collapse = ", "), character(1)),
  stringsAsFactors = FALSE
) %>%
  arrange(stage, desc(n_cells))

kable(sample_summary, caption = "Per-sample summary prior to integration")
```

```{r input_stage_counts, echo=TRUE}
if (!params$stage_col %in% colnames(srt_obj@meta.data)) {
  stop("Stage column not found in merged object metadata: ", params$stage_col)
}

stage_counts <- srt_obj@meta.data %>%
  count(.data[[params$stage_col]], name = "n_cells") %>%
  arrange(desc(n_cells))

kable(stage_counts, caption = "Cell counts by stage in merged SCT object")
```

## Integration by stage

Now we integrate each stage individually to evaluate how stage-specific replicates group after SCT-based integration.

```{r select_integration_features, echo=TRUE}
features_file <- file.path(params$out_dir, "srt_list_integration_features.rds")

if (file.exists(features_file)) {
  features <- readRDS(features_file)
  cat("Reusing previously saved integration features:\n")
  cat(features_file, "\n")
} else {
  features <- SelectIntegrationFeatures(
    object.list = sample_list,
    nfeatures = params$nfeatures
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(features, file = features_file)
  }
}

feature_summary <- data.frame(
  metric = c("n_integration_features"),
  value = c(length(features))
)

kable(feature_summary, caption = "Number of integration features selected across all samples")
```

```{r preprocess_for_integration, echo=TRUE}
sample_list <- lapply(
  X = sample_list,
  FUN = function(x) {
    x <- RunPCA(x, verbose = FALSE, features = features)
    return(x)
  }
)

gc()
```

```{r integrate_stages, echo=TRUE}
stage_list_file <- file.path(params$out_dir, "srt_list_integrated.rds")

if (file.exists(stage_list_file)) {
  stage_obj_list <- readRDS(stage_list_file)
  cat("Reusing previously saved stage-integrated object list:\n")
  cat(stage_list_file, "\n")
} else {
  stage_obj_list <- list()

  for (stage_id in params$stages) {
    cat("Integrating stage:", stage_id, "\n")

    filtered_list <- Filter(
      function(obj) {
        md <- obj@meta.data
        if (!params$stage_col %in% colnames(md)) return(FALSE)
        as.character(md[[params$stage_col]][1]) == stage_id
      },
      sample_list
    )

    if (length(filtered_list) == 0) {
      warning("No samples found for stage: ", stage_id, ". Skipping.")
      next
    }

    stage_name <- str_replace(stage_id, pattern = "-", replacement = "_")
    stage_file <- file.path(params$out_dir, paste0("srt_", stage_name, "_integrated.rds"))

    if (file.exists(stage_file)) {
      cat("Reusing previously saved integrated object for stage ", stage_id, ":\n", stage_file, "\n", sep = "")
      atlas_int <- readRDS(stage_file)
      stage_obj_list[[stage_name]] <- atlas_int
      next
    }

    options(future.globals.maxSize = 300 * 1024^3)

    filtered_list <- PrepSCTIntegration(
      object.list = filtered_list,
      anchor.features = features,
      verbose = FALSE
    )

    gc()

    anchors <- FindIntegrationAnchors(
      object.list = filtered_list,
      normalization.method = "SCT",
      anchor.features = features,
      reduction = "rpca",
      dims = 1:params$dims_use
    )

    gc()

    atlas_int <- IntegrateData(
      anchorset = anchors,
      normalization.method = "SCT",
      dims = 1:params$dims_use
    )

    gc()

    DefaultAssay(atlas_int) <- "integrated"
    atlas_int <- RunPCA(atlas_int, npcs = params$dims_use, verbose = FALSE)
    atlas_int <- RunUMAP(atlas_int, reduction = "pca", dims = 1:params$dims_use)
    atlas_int <- FindNeighbors(atlas_int, reduction = "pca", dims = 1:params$dims_use)
    atlas_int <- FindClusters(
      atlas_int,
      resolution = params$clustering_resolutions,
      algorithm = 4
    )

    stage_obj_list[[stage_name]] <- atlas_int

    if (isTRUE(params$save_intermediate_files)) {
      saveRDS(atlas_int, file = stage_file)
    }

    gc()
  }

  if (length(stage_obj_list) == 0) {
    stop("No stage-specific integrations were produced.")
  }

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(stage_obj_list, file = stage_list_file)
  }
}

if (length(stage_obj_list) == 0) {
  stop("No stage-specific integrations were produced.")
}

stage_summary_df <- bind_rows(lapply(names(stage_obj_list), function(stage_name) {
  obj <- stage_obj_list[[stage_name]]
  md <- obj@meta.data
  stage_value <- if (params$stage_col %in% colnames(md)) as.character(md[[params$stage_col]][1]) else stage_name
  data.frame(
    stage = stage_value,
    n_samples = if (params$sample_id_col %in% colnames(md)) length(unique(md[[params$sample_id_col]])) else NA_integer_,
    n_features = nrow(obj),
    n_cells = ncol(obj),
    stringsAsFactors = FALSE
  )
}))
```

### Stage integration summary

```{r stage_integration_summary_table, echo=TRUE}
kable(stage_summary_df, caption = "Summary of stage-specific integration outputs")
```

### Stage-specific UMAPs

```{r stage_specific_umaps, echo=TRUE, fig.width=12, fig.height=4.5}
stage_umap_plots <- lapply(names(stage_obj_list), function(stage_name) {
  obj <- stage_obj_list[[stage_name]]
  DimPlot(obj, reduction = "umap", group.by = params$sample_id_col) +
    ggtitle(paste0("Stage integration: ", stage_name))
})

wrap_plots(stage_umap_plots)
```

### Stage-specific UMAPs colored by clusters

```{r stage_specific_cluster_umaps, echo=TRUE, fig.width=12, fig.height=4.5}
stage_cluster_plots <- lapply(names(stage_obj_list), function(stage_name) {
  obj <- stage_obj_list[[stage_name]]
  default_cluster_col <- grep("_snn_res\\.", colnames(obj@meta.data), value = TRUE)[1]
  if (is.na(default_cluster_col) || is.null(default_cluster_col)) {
    return(ggplot() + ggtitle(paste0(stage_name, ": no cluster column found")))
  }
  DimPlot(obj, reduction = "umap", group.by = default_cluster_col, label = TRUE, repel = TRUE) +
    ggtitle(paste0("Clusters: ", stage_name))
})

wrap_plots(stage_cluster_plots)
```

## Integrate all samples using adult references

```{r adult_reference_summary, echo=TRUE}
adult_idx <- which(names(sample_list) %in% params$adult_reference_samples)

adult_ref_table <- data.frame(
  sample_id = names(sample_list),
  is_reference = names(sample_list) %in% params$adult_reference_samples,
  stringsAsFactors = FALSE
)

kable(adult_ref_table, caption = "Adult samples selected as references for all-sample integration")

if (length(adult_idx) == 0) {
  stop("No adult reference samples found in sample_list. Check params$adult_reference_samples.")
}
```

```{r prep_all_samples_for_anchor_integration, echo=TRUE}
sample_list_all <- lapply(
  X = sample_list,
  FUN = function(x) {
    x <- RunPCA(x, verbose = FALSE)
    return(x)
  }
)

options(future.globals.maxSize = 300 * 1024^3)

sample_list_all <- PrepSCTIntegration(
  object.list = sample_list_all,
  anchor.features = features,
  verbose = FALSE
)

gc()
```

```{r anchor_integration, echo=TRUE}
out_file_name <- file.path(params$out_dir, "srt_integrated_all.rds")

if (file.exists(out_file_name)) {
  atlas_int <- readRDS(out_file_name)
  cat("Reusing previously saved adult-anchored integrated atlas:\n")
  cat(out_file_name, "\n")
} else {
  anchors <- FindIntegrationAnchors(
    object.list = sample_list_all,
    normalization.method = "SCT",
    anchor.features = features,
    reference = adult_idx,
    reduction = "rpca",
    dims = 1:params$dims_use
  )

  gc()

  atlas_int <- IntegrateData(
    anchorset = anchors,
    normalization.method = "SCT",
    dims = 1:params$dims_use
  )

  gc()

  DefaultAssay(atlas_int) <- "integrated"
  atlas_int <- RunPCA(atlas_int, npcs = params$dims_use, verbose = FALSE)
  atlas_int <- RunUMAP(atlas_int, reduction = "pca", dims = 1:params$dims_use)
  atlas_int <- FindNeighbors(atlas_int, reduction = "pca", dims = 1:params$dims_use)
  atlas_int <- FindClusters(
    atlas_int,
    resolution = params$clustering_resolutions,
    algorithm = 4
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(atlas_int, file = out_file_name)
  }

  gc()
}
```

### Integrated atlas summary

```{r integrated_atlas_summary, echo=TRUE}
atlas_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay", "assays"),
  value = c(
    nrow(atlas_int),
    ncol(atlas_int),
    DefaultAssay(atlas_int),
    paste(Assays(atlas_int), collapse = ", ")
  ),
  stringsAsFactors = FALSE
)

kable(atlas_summary, caption = "Summary of the adult-anchored integrated atlas")
```

### Integrated atlas UMAP by stage and sample

```{r integrated_atlas_umaps, echo=TRUE, fig.width=12, fig.height=4.5}
p_stage <- DimPlot(atlas_int, reduction = "umap", group.by = params$stage_col) +
  ggtitle("Integrated atlas colored by stage")

p_sample <- DimPlot(atlas_int, reduction = "umap", group.by = params$sample_id_col) +
  ggtitle("Integrated atlas colored by sample")

p_stage + p_sample
```

### Integrated atlas clusters

```{r integrated_atlas_clusters, echo=TRUE, fig.width=7, fig.height=5}
default_cluster_col <- grep("_snn_res\\.", colnames(atlas_int@meta.data), value = TRUE)[1]

if (!is.na(default_cluster_col) && !is.null(default_cluster_col)) {
  DimPlot(atlas_int, reduction = "umap", group.by = default_cluster_col, label = TRUE, repel = TRUE) +
    ggtitle(paste0("Integrated atlas clusters (", default_cluster_col, ")"))
}
```

### Cluster counts

```{r integrated_cluster_counts, echo=TRUE}
if (!is.na(default_cluster_col) && !is.null(default_cluster_col)) {
  cluster_counts <- atlas_int@meta.data %>%
    count(.data[[default_cluster_col]], name = "n_cells") %>%
    arrange(desc(n_cells))

  kable(cluster_counts, caption = paste0("Cluster sizes for ", default_cluster_col))
}
```

### Output files

```{r output_files_summary, echo=TRUE}
stage_files <- file.path(
  params$out_dir,
  paste0("srt_", gsub("-", "_", params$stages), "_integrated.rds")
)

output_files <- data.frame(
  object = c(
    "List of stage-integrated objects",
    "Integration features",
    "Adult-anchored integrated atlas",
    paste0("Stage object: ", gsub("-", "_", params$stages))
  ),
  file = c(
    file.path(params$out_dir, "srt_list_integrated.rds"),
    file.path(params$out_dir, "srt_list_integration_features.rds"),
    file.path(params$out_dir, "srt_integrated_all.rds"),
    stage_files
  ),
  exists = file.exists(c(
    file.path(params$out_dir, "srt_list_integrated.rds"),
    file.path(params$out_dir, "srt_list_integration_features.rds"),
    file.path(params$out_dir, "srt_integrated_all.rds"),
    stage_files
  )),
  stringsAsFactors = FALSE
)

kable(output_files, caption = "Output files produced by this report")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
