---
title: "04_IntegrateSeuratObjects"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  sct_list_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/sct_list.rds"
  merged_sct_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_obj_norm_scaled_RNA_plus_SCT.rds"
  sample_color_table: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas/data/metadata/sample_color_codes.tsv"
  stage_color_table: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas/data/metadata/stage_color_codes.tsv"
  sample_id_col: "sample_id"
  stage_col: "Stage"
  cluster_col: "integrated_snn_res.1.8"
  stages: !r c("0_1_days", "3_days", "7_days", "Adult")
  adult_reference_samples: !r c("2_Adult_Unsorted", "2_Adult_Unsorted_DASH", "7_Sorted_Adult")
  nfeatures: 3000
  dims_use: 40
  clustering_resolutions: !r c(0.6, 0.8, 1.0, 1.2, 1.4, 1.6, 1.8, 2.0, 2.2, 2.4, 2.6, 2.8, 3)
  save_intermediate_files: TRUE
  save_figures: TRUE
  figure_dpi: 600
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"
  seed: 42
---

# Description

Integrate Seurat objects.

This template is designed to work with the outputs of the previous step (**03_ComputeSCT.Rmd**) and performs two integration procedures:

- integrate replicates within each stage independently
- integrate all samples using adult samples as references

The report also provides diagnostic tables and figures to summarize inputs and integration outputs. Whenever intermediate files already exist, the report reuses them instead of recomputing from scratch.

Figures are displayed in the HTML report and also saved as publication-quality PDF and PNG files. Sample colors and clean sample names are read from an external metadata table.

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(sctransform)
library(scCustomize)
library(DT)
library(stringr)
library(ggplot2)
library(dplyr)
library(knitr)
library(patchwork)
library(leidenbase)
```

### Parameters

Parameters used to generate this report:

```{r print_parameters, echo=TRUE}
str(params)
```

### Load helper functions

```{r load_functions, echo=TRUE}
bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R",
    "umap_helpers.R"
  )
)
```

### Plotting helpers

```{r plotting_helpers, echo=TRUE}
save_publication_plot <- function(plot_obj,
                                  file_stub,
                                  out_dir,
                                  width = 10,
                                  height = 6,
                                  dpi = 600) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  ggsave(
    filename = file.path(out_dir, paste0(file_stub, ".pdf")),
    plot = plot_obj,
    width = width,
    height = height,
    units = "in",
    dpi = dpi,
    bg = "white"
  )
  
  ggsave(
    filename = file.path(out_dir, paste0(file_stub, ".png")),
    plot = plot_obj,
    width = width,
    height = height,
    units = "in",
    dpi = dpi,
    bg = "white"
  )
}

clean_sample_metadata <- function(obj, sample_id_col, sample_key) {
  if (!sample_id_col %in% colnames(obj@meta.data)) {
    stop("sample_id_col not found in object metadata: ", sample_id_col)
  }
  
  obj@meta.data[[sample_id_col]] <- as.character(obj@meta.data[[sample_id_col]])
  
  obj@meta.data$CleanSampleID <- dplyr::recode(
    obj@meta.data[[sample_id_col]],
    !!!setNames(sample_key$CleanSampleID, sample_key$RawSampleID),
    .default = obj@meta.data[[sample_id_col]]
  )
  
  obj
}

clean_stage_metadata <- function(obj, stage_id_col, stage_key) {
  if (!stage_id_col %in% colnames(obj@meta.data)) {
    stop("stage_id_col not found in object metadata: ", stage_id_col)
  }
  
  obj@meta.data[[stage_id_col]] <- as.character(obj@meta.data[[stage_id_col]])
  
  obj@meta.data$CleanStageID <- dplyr::recode(
    obj@meta.data[[stage_id_col]],
    !!!setNames(stage_key$CleanStageID, stage_key$RawStageID),
    .default = obj@meta.data[[stage_id_col]]
  )
  
  obj
}

make_named_sample_colors <- function(sample_key) {
  setNames(sample_key$HexCode, sample_key$CleanSampleID)
}

make_named_stage_colors <- function(stage_key) {
  setNames(stage_key$HexCode, stage_key$StageID)
}

```

### Prepare output directory

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)
figure_dir <- file.path(params$out_dir, "integration_figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n\n")

cat("Figure directory:\n")
cat(normalizePath(figure_dir, mustWork = TRUE), "\n")
```

## Load sample color table

```{r load_sample_colors, echo=TRUE}
if (!file.exists(path.expand(params$sample_color_table))) {
  stop("sample_color_table not found: ", params$sample_color_table)
}

sample_key <- read.table(
  file = path.expand(params$sample_color_table),
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE,
  comment.char = "",
  check.names = FALSE
)

required_color_cols <- c("RawSampleID", "CleanSampleID", "HexCode")
missing_color_cols <- setdiff(required_color_cols, colnames(sample_key))

if (length(missing_color_cols) > 0) {
  stop("Missing required columns in sample_color_table: ",
       paste(missing_color_cols, collapse = ", "))
}

sample_key <- sample_key %>%
  distinct(RawSampleID, .keep_all = TRUE)

sample_colors <- make_named_sample_colors(sample_key)

kable(sample_key, caption = "Sample color and clean-label table used for plotting")
```

## Load stage color table

```{r load_stage_colors, echo=TRUE}
if (!file.exists(path.expand(params$stage_color_table))) {
  stop("stage_color_table not found: ", params$stage_color_table)
}

stage_key <- read.table(
  file = path.expand(params$stage_color_table),
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE,
  comment.char = "",
  check.names = FALSE
)

required_color_cols <- c("RawStageID", "CleanStageID", "HexCode")
missing_color_cols <- setdiff(required_color_cols, colnames(stage_key))

if (length(missing_color_cols) > 0) {
  stop("Missing required columns in stage_color_table: ",
       paste(missing_color_cols, collapse = ", "))
}

stage_key <- stage_key %>%
  distinct(RawStageID, .keep_all = TRUE)

stage_colors <- make_named_stage_colors(stage_key)

kable(stage_key, caption = "Stage colors and labels used for plotting")
```

## Load data

```{r load_data, echo=TRUE}
if (!file.exists(params$sct_list_file)) {
  stop("sct_list_file not found: ", params$sct_list_file)
}
if (!file.exists(params$merged_sct_file)) {
  stop("merged_sct_file not found: ", params$merged_sct_file)
}

sample_list <- readRDS(file = params$sct_list_file)
srt_obj <- readRDS(file = params$merged_sct_file)

gc()

cat("Loaded sample_list with", length(sample_list), "objects.\n")
cat("Merged object dimensions (features x cells):\n")
print(dim(srt_obj))
```

## Input summaries

```{r input_object_summary, echo=TRUE}
input_summary <- data.frame(
  metric = c("n_samples", "merged_n_features", "merged_n_cells", "merged_assays"),
  value = c(
    length(sample_list),
    nrow(srt_obj),
    ncol(srt_obj),
    paste(Assays(srt_obj), collapse = ", ")
  ),
  stringsAsFactors = FALSE
)

kable(input_summary, caption = "Summary of the SCT inputs used for integration")
```

```{r input_sample_table, echo=TRUE}
sample_summary <- data.frame(
  sample_id = names(sample_list),
  n_features = vapply(sample_list, nrow, numeric(1)),
  n_cells = vapply(sample_list, ncol, numeric(1)),
  stage = vapply(
    sample_list,
    function(x) {
      md <- x@meta.data
      if (params$stage_col %in% colnames(md)) as.character(md[[params$stage_col]][1]) else NA_character_
    },
    character(1)
  ),
  clean_sample_id = vapply(
    sample_list,
    function(x) {
      md <- x@meta.data
      if ("CleanSampleID" %in% colnames(md)) as.character(md$CleanSampleID[1]) else NA_character_
    },
    character(1)
  ),
  assays = vapply(sample_list, function(x) paste(Assays(x), collapse = ", "), character(1)),
  stringsAsFactors = FALSE
) %>%
  arrange(stage, desc(n_cells))

kable(sample_summary, caption = "Per-sample summary prior to integration")
```

```{r input_stage_counts, echo=TRUE}
if (!params$stage_col %in% colnames(srt_obj@meta.data)) {
  stop("Stage column not found in merged object metadata: ", params$stage_col)
}

stage_counts <- srt_obj@meta.data %>%
  count(.data[[params$stage_col]], name = "n_cells") %>%
  arrange(desc(n_cells))

kable(stage_counts, caption = "Cell counts by stage in merged SCT object")
```

## Integration by stage

Now we integrate each stage individually to evaluate how stage-specific replicates group after SCT-based integration.

```{r select_integration_features, echo=TRUE}
features_file <- file.path(params$out_dir, "srt_list_integration_features.rds")

if (file.exists(features_file)) {
  features <- readRDS(features_file)
  cat("Reusing previously saved integration features:\n")
  cat(features_file, "\n")
} else {
  features <- SelectIntegrationFeatures(
    object.list = sample_list,
    nfeatures = params$nfeatures
  )
  
  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(features, file = features_file)
  }
}

feature_summary <- data.frame(
  metric = c("n_integration_features"),
  value = c(length(features))
)

kable(feature_summary, caption = "Number of integration features selected across all samples")
```

```{r preprocess_for_integration, echo=TRUE}

# get PCs for each object as described in https://satijalab.org/seurat/articles/integration_rpca.html

sample_list <- lapply(
  X = sample_list,
  FUN = function(x) {
    x <- RunPCA(x, verbose = FALSE, features = features)
    return(x)
  }
)

gc()
```

```{r integrate_stages, echo=TRUE}
stage_list_file <- file.path(params$out_dir, "srt_list_integrated.rds")

if (file.exists(stage_list_file)) {
  stage_obj_list <- readRDS(stage_list_file)
  cat("Reusing previously saved stage-integrated object list:\n")
  cat(stage_list_file, "\n")
} else {
  stage_obj_list <- list()
  
  for (stage_id in params$stages) {
    cat("Integrating stage:", stage_id, "\n")
    
    filtered_list <- Filter(
      function(obj) {
        md <- obj@meta.data
        if (!params$stage_col %in% colnames(md)) return(FALSE)
        as.character(md[[params$stage_col]][1]) == stage_id
      },
      sample_list
    )
    
    if (length(filtered_list) == 0) {
      warning("No samples found for stage: ", stage_id, ". Skipping.")
      next
    }
    
    stage_name <- str_replace(stage_id, pattern = "-", replacement = "_")
    stage_file <- file.path(params$out_dir, paste0("srt_", stage_name, "_integrated.rds"))
    
    if (file.exists(stage_file)) {
      cat("Reusing previously saved integrated object for stage ", stage_id, ":\n", stage_file, "\n", sep = "")
      atlas_tmp <- readRDS(stage_file)
      stage_obj_list[[stage_name]] <- atlas_tmp
      next
    }
    
    options(future.globals.maxSize = 300 * 1024^3)
    
    filtered_list <- PrepSCTIntegration(
      object.list = filtered_list,
      anchor.features = features,
      verbose = FALSE
    )
    
    gc()
    
    anchors <- FindIntegrationAnchors(
      object.list = filtered_list,
      normalization.method = "SCT",
      anchor.features = features,
      reduction = "rpca",
      dims = 1:params$dims_use
    )
    
    gc()
    
    atlas_tmp <- IntegrateData(
      anchorset = anchors,
      normalization.method = "SCT",
      dims = 1:params$dims_use
    )
    
    gc()
    
    DefaultAssay(atlas_tmp) <- "integrated"
    atlas_tmp <- RunPCA(atlas_tmp, npcs = params$dims_use, verbose = FALSE)
    atlas_tmp <- RunUMAP(atlas_tmp, reduction = "pca", dims = 1:params$dims_use)
    atlas_tmp <- FindNeighbors(atlas_tmp, reduction = "pca", dims = 1:params$dims_use)
    atlas_tmp <- FindClusters(
      atlas_tmp,
      resolution = params$clustering_resolutions,
      algorithm = 4
    )
    
    atlas_tmp <- clean_sample_metadata(
      obj = atlas_tmp,
      sample_id_col = params$sample_id_col,
      sample_key = sample_key
    )
    
    atlas_tmp <- clean_stage_metadata(
      obj = atlas_tmp,
      stage_id_col = params$stage_col,
      stage_key = stage_key
    )
    
    stage_obj_list[[stage_name]] <- atlas_tmp
    
    if (isTRUE(params$save_intermediate_files)) {
      saveRDS(atlas_tmp, file = stage_file)
    }
    
    gc()
  }
  
  if (length(stage_obj_list) == 0) {
    stop("No stage-specific integrations were produced.")
  }
  
  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(stage_obj_list, file = stage_list_file)
  }
}

if (length(stage_obj_list) == 0) {
  stop("No stage-specific integrations were produced.")
}

stage_obj_list <- lapply(stage_obj_list, clean_sample_metadata,
                         sample_id_col = params$sample_id_col,
                         sample_key = sample_key)

stage_summary_df <- bind_rows(lapply(names(stage_obj_list), function(stage_name) {
  obj <- stage_obj_list[[stage_name]]
  md <- obj@meta.data
  stage_value <- if (params$stage_col %in% colnames(md)) as.character(md[[params$stage_col]][1]) else stage_name
  data.frame(
    stage = stage_value,
    n_samples = if (params$sample_id_col %in% colnames(md)) length(unique(md[[params$sample_id_col]])) else NA_integer_,
    n_features = nrow(obj),
    n_cells = ncol(obj),
    stringsAsFactors = FALSE
  )
}))
```

### Stage integration summary

```{r stage_integration_summary_table, echo=TRUE}
kable(stage_summary_df, caption = "Summary of stage-specific integration outputs")
```

### Stage-specific UMAPs

```{r stage_specific_umaps, echo=TRUE, fig.width=12, fig.height=4.5}

stage_umap_plots <- lapply(names(stage_obj_list), function(stage_name) {
  obj <- stage_obj_list[[stage_name]]
  plot_title <- stage_key$CleanStageID[stage_key$RawStageID == stage_name]
  
  p <- DimPlot(
    obj,
    reduction = "umap",
    group.by = "CleanSampleID",
    cols = sample_colors,
    shuffle = TRUE,
  ) +
    #    ggtitle(paste0("Stage integration: ", stage_name)) +
    theme_no_axes +
    # theme_classic(base_size = 11) +
    # theme(
    #   plot.title = element_text(face = "bold", hjust = 0.5),
    #   legend.title = element_blank()
    # ) +
    coord_equal()
  
  p <- add_umap_compass(p, hjust_x = -0.02)
  p <- p + ggtitle(plot_title)
  
  if (isTRUE(params$save_figures)) {
    save_publication_plot(
      plot_obj = p,
      file_stub = paste0("stage_umap_by_sample_", stage_name),
      out_dir = figure_dir,
      width = 8,
      height = 5,
      dpi = params$figure_dpi
    )
  }
  
  p
})

wrap_plots(stage_umap_plots)
```

### Stage-specific UMAPs colored by clusters

```{r stage_specific_cluster_umaps, echo=TRUE, fig.width=12, fig.height=4.5}
stage_cluster_plots <- lapply(names(stage_obj_list), function(stage_name) {
  obj <- stage_obj_list[[stage_name]]
  default_cluster_col <- params$cluster_col
  plot_title <- stage_key$CleanStageID[stage_key$RawStageID == stage_name]
  if (is.na(default_cluster_col) || is.null(default_cluster_col)) {
    return(ggplot() + ggtitle(paste0(plot_title, ": no cluster column found")))
  }
  
  p <- DimPlot(obj, reduction = "umap", group.by = default_cluster_col, label = TRUE, repel = TRUE) +
    ggtitle(paste0("Clusters: ", plot_title)) +
    #    theme_classic(base_size = 11) +
    #    theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
    coord_equal() +
    theme_no_axes
  
  p <- add_umap_compass(p, hjust_x = -0.02)
  
  if (isTRUE(params$save_figures)) {
    save_publication_plot(
      plot_obj = p,
      file_stub = paste0("stage_umap_by_cluster_", stage_name),
      out_dir = figure_dir,
      width = 8,
      height = 5,
      dpi = params$figure_dpi
    )
  }
  
  p
})

wrap_plots(stage_cluster_plots)
```

## Integrate all samples using adult references

```{r adult_reference_summary, echo=TRUE}
adult_idx <- which(names(sample_list) %in% params$adult_reference_samples)

adult_ref_table <- data.frame(
  sample_id = names(sample_list),
  clean_sample_id = dplyr::recode(
    names(sample_list),
    !!!setNames(sample_key$CleanSampleID, sample_key$RawSampleID),
    .default = names(sample_list)
  ),
  is_reference = names(sample_list) %in% params$adult_reference_samples,
  stringsAsFactors = FALSE
)

kable(adult_ref_table, caption = "Adult samples selected as references for all-sample integration")

if (length(adult_idx) == 0) {
  stop("No adult reference samples found in sample_list. Check params$adult_reference_samples.")
}
```

```{r prep_all_samples_for_anchor_integration, echo=TRUE}
sample_list_all <- lapply(
  X = sample_list,
  FUN = function(x) {
    x <- RunPCA(x, verbose = FALSE, features = features)
    return(x)
  }
)

options(future.globals.maxSize = 300 * 1024^3)

sample_list_all <- PrepSCTIntegration(
  object.list = sample_list_all,
  anchor.features = features,
  verbose = FALSE
)

gc()
```

```{r anchor_integration, echo=TRUE}
out_file_name <- file.path(params$out_dir, "srt_integrated_all.rds")

if (file.exists(out_file_name)) {
  atlas_int <- readRDS(out_file_name)
  cat("Reusing previously saved adult-anchored integrated atlas:\n")
  cat(out_file_name, "\n")
} else {
  anchors <- FindIntegrationAnchors(
    object.list = sample_list_all,
    normalization.method = "SCT",
    anchor.features = features,
    reference = adult_idx,
    reduction = "rpca",
    dims = 1:params$dims_use
  )
  
  gc()
  
  atlas_int <- IntegrateData(
    anchorset = anchors,
    normalization.method = "SCT",
    dims = 1:params$dims_use
  )
  
  gc()
  
  DefaultAssay(atlas_int) <- "integrated"
  atlas_int <- RunPCA(atlas_int, npcs = params$dims_use, verbose = FALSE)
  atlas_int <- RunUMAP(atlas_int, reduction = "pca", dims = 1:params$dims_use)
  atlas_int <- FindNeighbors(atlas_int, reduction = "pca", dims = 1:params$dims_use)
  atlas_int <- FindClusters(
    atlas_int,
    resolution = params$clustering_resolutions,
    algorithm = 4
  )
  
  atlas_int <- clean_sample_metadata(
    obj = atlas_int,
    sample_id_col = params$sample_id_col,
    sample_key = sample_key
  )
  
  atlas_int <- clean_stage_metadata(
    obj = atlas_int,
    stage_id_col = params$stage_col,
    stage_key = stage_key
  )
  
  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(atlas_int, file = out_file_name)
  }
  
  gc()
}


```

### Integrated atlas summary

```{r integrated_atlas_summary, echo=TRUE}
atlas_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay", "assays"),
  value = c(
    nrow(atlas_int),
    ncol(atlas_int),
    DefaultAssay(atlas_int),
    paste(Assays(atlas_int), collapse = ", ")
  ),
  stringsAsFactors = FALSE
)

kable(atlas_summary, caption = "Summary of the adult-anchored integrated atlas")
```

### Integrated atlas UMAP by stage and sample

```{r integrated_atlas_umaps, echo=TRUE, fig.width=12, fig.height=4.5}

p_stage <- DimPlot(atlas_int,
                   reduction = "umap",
                   group.by = "CleanStageID",
                   cols = stage_colors,
                   shuffle = TRUE) +
  ggtitle("Integrated atlas - stage") +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  coord_equal() +
  theme_no_axes

p_stage <- add_umap_compass(p_stage, hjust_x = -0.02)

p_sample <- DimPlot(atlas_int,
                    reduction = "umap",
                    group.by = "CleanSampleID",
                    cols = sample_colors,
                    shuffle = TRUE) +
  ggtitle("Integrated atlas - samples") +
  theme_classic(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.title = element_blank()
  ) +
  coord_equal()

p_sample <- add_umap_compass(p_sample, hjust_x = -0.02)

if (isTRUE(params$save_figures)) {
  save_publication_plot(
    plot_obj = p_stage,
    file_stub = "integrated_atlas_umap_by_stage",
    out_dir = figure_dir,
    width = 8,
    height = 5,
    dpi = params$figure_dpi
  )
  save_publication_plot(
    plot_obj = p_sample,
    file_stub = "integrated_atlas_umap_by_sample",
    out_dir = figure_dir,
    width = 8,
    height = 5,
    dpi = params$figure_dpi
  )
}

p_stage + p_sample
```

### Integrated atlas clusters

```{r integrated_atlas_clusters, echo=TRUE, fig.width=7, fig.height=5}
default_cluster_col <- params$cluster_col

if (!is.na(default_cluster_col) && !is.null(default_cluster_col)) {
  p_clusters <- DimPlot(atlas_int,
                        reduction = "umap",
                        group.by = default_cluster_col,
                        label = TRUE,
                        repel = TRUE) +
    ggtitle(paste0("Integrated atlas clusters (", default_cluster_col, ")")) +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
    coord_equal() +
    theme_no_axes
  
  p_clusters <- add_umap_compass(p_clusters, hjust_x = -0.02)
  
  if (isTRUE(params$save_figures)) {
    save_publication_plot(
      plot_obj = p_clusters,
      file_stub = paste0("integrated_atlas_clusters_", gsub("\\.", "_", default_cluster_col)),
      out_dir = figure_dir,
      width = 8,
      height = 5,
      dpi = params$figure_dpi
    )
  }
  
  p_clusters
}
```

### Cluster counts

```{r integrated_cluster_counts, echo=TRUE}
if (!is.na(default_cluster_col) && !is.null(default_cluster_col)) {
  cluster_counts <- atlas_int@meta.data %>%
    count(.data[[default_cluster_col]], name = "n_cells") %>%
    arrange(desc(n_cells))
  
  kable(cluster_counts, caption = paste0("Cluster sizes for ", default_cluster_col))
}
```

### Output files

```{r output_files_summary, echo=TRUE}
stage_files <- file.path(
  params$out_dir,
  paste0("srt_", gsub("-", "_", params$stages), "_integrated.rds")
)

output_files <- data.frame(
  object = c(
    "List of stage-integrated objects",
    "Integration features",
    "Adult-anchored integrated atlas",
    paste0("Stage object: ", gsub("-", "_", params$stages))
  ),
  file = c(
    file.path(params$out_dir, "srt_list_integrated.rds"),
    file.path(params$out_dir, "srt_list_integration_features.rds"),
    file.path(params$out_dir, "srt_integrated_all.rds"),
    stage_files
  ),
  exists = file.exists(c(
    file.path(params$out_dir, "srt_list_integrated.rds"),
    file.path(params$out_dir, "srt_list_integration_features.rds"),
    file.path(params$out_dir, "srt_integrated_all.rds"),
    stage_files
  )),
  stringsAsFactors = FALSE
)

kable(output_files, caption = "Output files produced by this report")
```

```{r figure_file_summary, echo=TRUE}
saved_figures <- list.files(figure_dir, full.names = FALSE)

figure_table <- data.frame(
  figure_file = saved_figures,
  stringsAsFactors = FALSE
)

kable(figure_table, caption = "Publication-quality integration figures saved to disk")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
