---
title: "05_ComputeModuleScores"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  stage_integrated_list_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_list_integrated.rds"
  integrated_atlas_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/srt_integrated_all.rds"
  marker_gene_table: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas/data/metadata/literature_marker_genes.tsv"
  sample_id_col: "sample_id"
  stage_col: "Stage"
  save_intermediate_files: TRUE
  save_figures: TRUE
  figure_dpi: 600
  dotplot_resolution: 1.8
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/"
  palette: ""
  seed: 42
---

# Description

Add module scores for major cell types to:

- the stage-specific integrated Seurat objects generated in the previous step
- the adult-anchored integrated Seurat object

This report also generates summary tables and publication-ready figures, including:

- feature plots for module scores
- dot plots for module scores

If cached objects with module scores already exist, they are loaded instead of recomputing them from scratch.

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(sctransform)
library(scCustomize)
library(DT)
library(stringr)
library(ggplot2)
library(dplyr)
library(knitr)
library(patchwork)
```

### Parameters

Parameters used to generate this report:

```{r print_parameters, echo=TRUE}
str(params)
```

### Load helper functions

```{r load_functions, echo=TRUE}
bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R",
    "module_score_helpers.R"
  )
)
```

### Prepare output directories

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)

figure_dir <- file.path(params$out_dir, "module_score_figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n\n")

cat("Figure directory:\n")
cat(normalizePath(figure_dir, mustWork = TRUE), "\n")
```

## Load integrated objects

```{r load_objects, echo=TRUE}
if (!file.exists(params$stage_integrated_list_file)) {
  stop("stage_integrated_list_file not found: ", params$stage_integrated_list_file)
}
if (!file.exists(params$integrated_atlas_file)) {
  stop("integrated_atlas_file not found: ", params$integrated_atlas_file)
}

stage_obj_list <- readRDS(params$stage_integrated_list_file)
atlas_int <- readRDS(params$integrated_atlas_file)

cat("Loaded", length(stage_obj_list), "stage-specific integrated objects.\n")
cat("Integrated atlas dimensions (features x cells):\n")
print(dim(atlas_int))
```

## Import marker genes

```{r import_module_score_table, echo=TRUE}
marker_genes <- read_marker_table(params$marker_gene_table)

datatable(marker_genes, options = list(pageLength = 10, scrollX = TRUE))
```

```{r marker_summary_table, echo=TRUE}
marker_summary <- summarize_marker_table(marker_genes)

kable(marker_summary, caption = "Number of marker genes per literature-defined cell/tissue type")
```

```{r build_marker_lists, echo=TRUE}
marker_genes_list <- build_marker_gene_list(marker_genes)

marker_list_summary <- data.frame(
  module = names(marker_genes_list),
  n_marker_genes = vapply(marker_genes_list, length, integer(1)),
  stringsAsFactors = FALSE
) %>%
  arrange(desc(n_marker_genes), module)

kable(marker_list_summary, caption = "Module score gene lists used in this report")
```

## Marker overlap diagnostics

```{r overlap_integrated_atlas, echo=TRUE}
atlas_overlap <- module_score_overlap_table(
  obj = atlas_int,
  gene_lists = marker_genes_list,
  object_name = "integrated_atlas"
)

kable(atlas_overlap, caption = "Marker overlap between literature gene lists and the integrated atlas")
```

```{r overlap_stage_objects, echo=TRUE}
stage_overlap <- bind_rows(
  lapply(names(stage_obj_list), function(nm) {
    module_score_overlap_table(
      obj = stage_obj_list[[nm]],
      gene_lists = marker_genes_list,
      object_name = nm
    )
  })
)

kable(stage_overlap, caption = "Marker overlap between literature gene lists and each stage-specific integrated object")
```

## Add module scores

```{r add_module_scores, echo=TRUE}
stage_scores_file <- file.path(params$out_dir, "srt_list_integrated_module_scores.rds")
atlas_scores_file <- file.path(params$out_dir, "srt_integrated_all_module_scores.rds")

if (file.exists(stage_scores_file) && file.exists(atlas_scores_file)) {
  stage_obj_list <- readRDS(stage_scores_file)
  atlas_int <- readRDS(atlas_scores_file)
  cat("Reusing cached module score objects:\n")
  cat("  ", stage_scores_file, "\n")
  cat("  ", atlas_scores_file, "\n")
} else {
  stage_obj_list <- lapply(stage_obj_list, function(obj) {
    add_module_scores_named(
      obj = obj,
      gene_lists = marker_genes_list,
      assay = "RNA",
      prefix = "MS_",
      seed = params$seed
    )
  })

  atlas_int <- add_module_scores_named(
    obj = atlas_int,
    gene_lists = marker_genes_list,
    assay = "RNA",
    prefix = "MS_",
    seed = params$seed
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(stage_obj_list, file = stage_scores_file)
    saveRDS(atlas_int, file = atlas_scores_file)
  }
}
```

```{r module_score_columns, echo=TRUE}
module_score_cols <- grep("^MS_", colnames(atlas_int@meta.data), value = TRUE)

module_score_table <- data.frame(
  module_score_column = module_score_cols,
  stringsAsFactors = FALSE
)

kable(module_score_table, caption = "Module score columns added to the Seurat metadata")
```

## Stage-specific module score summaries

```{r stage_module_score_summary, echo=TRUE}
stage_module_summary <- bind_rows(lapply(names(stage_obj_list), function(nm) {
  obj <- stage_obj_list[[nm]]
  cols <- grep("^MS_", colnames(obj@meta.data), value = TRUE)
  data.frame(
    stage_object = nm,
    n_module_score_columns = length(cols),
    n_cells = ncol(obj),
    n_features = nrow(obj),
    cluster_column = find_cluster_column(obj, params$dotplot_resolution),
    stringsAsFactors = FALSE
  )
}))

kable(stage_module_summary, caption = "Summary of stage-specific objects after adding module scores")
```

## Feature plots by stage

```{r stage_feature_plots, echo=TRUE, fig.width=14, fig.height=10}
stage_feature_plots <- lapply(names(stage_obj_list), function(nm) {
  obj <- stage_obj_list[[nm]]
  p <- plot_module_feature_grid(
    obj = obj,
    module_cols = module_score_cols,
    reduction = "umap",
    ncol = 5,
    pt.size = 0.05
  ) + patchwork::plot_annotation(title = paste0("Module scores: ", nm))

  if (isTRUE(params$save_figures)) {
    save_publication_plot(
      plot_obj = p,
      file_stub = paste0("featureplot_module_scores_", nm),
      out_dir = figure_dir,
      width = 16,
      height = 11,
      dpi = params$figure_dpi
    )
  }

  p
})

for (p in stage_feature_plots) print(p)
```

## Dot plots by stage

```{r stage_dot_plots, echo=TRUE, fig.width=12, fig.height=6}
stage_dot_plots <- lapply(names(stage_obj_list), function(nm) {
  obj <- stage_obj_list[[nm]]
  cluster_col <- find_cluster_column(obj, params$dotplot_resolution)

  if (is.na(cluster_col)) {
    return(ggplot() + ggtitle(paste0(nm, ": no cluster column found for dot plot")))
  }

  p <- plot_module_dotplot(
    obj = obj,
    module_cols = module_score_cols,
    group.by = cluster_col
  ) + ggtitle(paste0("Module score dot plot: ", nm, " (", cluster_col, ")"))

  if (isTRUE(params$save_figures)) {
    save_publication_plot(
      plot_obj = p,
      file_stub = paste0("dotplot_module_scores_", nm, "_", gsub("\\.", "_", cluster_col)),
      out_dir = figure_dir,
      width = 14,
      height = 7,
      dpi = params$figure_dpi
    )
  }

  p
})

for (p in stage_dot_plots) print(p)
```

## Integrated atlas module score figures

```{r integrated_atlas_feature_plot, echo=TRUE, fig.width=14, fig.height=10}
p_atlas_feature <- plot_module_feature_grid(
  obj = atlas_int,
  module_cols = module_score_cols,
  reduction = "umap",
  ncol = 5,
  pt.size = 0.05
) + patchwork::plot_annotation(title = "Integrated atlas module scores")

if (isTRUE(params$save_figures)) {
  save_publication_plot(
    plot_obj = p_atlas_feature,
    file_stub = "featureplot_module_scores_integrated_atlas",
    out_dir = figure_dir,
    width = 16,
    height = 11,
    dpi = params$figure_dpi
  )
}

p_atlas_feature
```

```{r integrated_atlas_dotplots, echo=TRUE, fig.width=12, fig.height=6}
atlas_cluster_col <- find_cluster_column(atlas_int, params$dotplot_resolution)

p_atlas_dot_cluster <- plot_module_dotplot(
  obj = atlas_int,
  module_cols = module_score_cols,
  group.by = atlas_cluster_col
) + ggtitle(paste0("Integrated atlas module score dot plot by cluster (", atlas_cluster_col, ")"))

p_atlas_dot_stage <- plot_module_dotplot(
  obj = atlas_int,
  module_cols = module_score_cols,
  group.by = params$stage_col
) + ggtitle("Integrated atlas module score dot plot by stage")

if (isTRUE(params$save_figures)) {
  save_publication_plot(
    plot_obj = p_atlas_dot_cluster,
    file_stub = paste0("dotplot_module_scores_integrated_atlas_", gsub("\\.", "_", atlas_cluster_col)),
    out_dir = figure_dir,
    width = 14,
    height = 7,
    dpi = params$figure_dpi
  )
  save_publication_plot(
    plot_obj = p_atlas_dot_stage,
    file_stub = "dotplot_module_scores_integrated_atlas_by_stage",
    out_dir = figure_dir,
    width = 14,
    height = 7,
    dpi = params$figure_dpi
  )
}

p_atlas_dot_cluster
p_atlas_dot_stage
```

## Integrated atlas summary

```{r integrated_atlas_summary, echo=TRUE}
atlas_summary <- data.frame(
  metric = c(
    "n_features",
    "n_cells",
    "default_assay",
    "cluster_column_for_dotplot",
    "n_module_score_columns"
  ),
  value = c(
    nrow(atlas_int),
    ncol(atlas_int),
    DefaultAssay(atlas_int),
    atlas_cluster_col,
    length(module_score_cols)
  ),
  stringsAsFactors = FALSE
)

kable(atlas_summary, caption = "Summary of the integrated atlas after adding module scores")
```

## Output files

```{r output_files_summary, echo=TRUE}
output_files <- data.frame(
  object = c(
    "Stage-integrated objects with module scores",
    "Adult-anchored integrated atlas with module scores"
  ),
  file = c(
    stage_scores_file,
    atlas_scores_file
  ),
  exists = file.exists(c(stage_scores_file, atlas_scores_file)),
  stringsAsFactors = FALSE
)

kable(output_files, caption = "Module score objects produced by this report")
```

```{r figure_file_summary, echo=TRUE}
saved_figures <- list.files(figure_dir, full.names = FALSE)

figure_table <- data.frame(
  figure_file = saved_figures,
  stringsAsFactors = FALSE
)

kable(figure_table, caption = "Publication-ready module score figures saved to disk")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```