---
title: "07_6_Subset_S_mediterranea_Neoblast_cells_to_H5AD"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  input_seurat_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mediterranea/S_mediterranea_seurat_annotated.rds"
  output_h5ad_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mediterranea/S_mediterranea_Neoblast_cells.h5ad"
  output_seurat_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mediterranea/S_mediterranea_Neoblast_cells.rds"
  annotation_col: "Major.cluster.description"
  clusters_to_keep:
    value:
      - "Neoblast"
  root_cluster_id: "Smed_neoblast_"
  cluster_id_col: "Subcluster.Number"
  final_label_col: "final_label_SAMap"

  assay: "RNA"
  layer: "counts"
  save_intermediate_files: TRUE
  seed: 42
---

# Description

Subset the annotated *Schmidtea mediterranea* reference atlas to cells annotated as neoblast cells and export the subset as an AnnData/H5AD object.

This report:

- loads the annotated *S. mediterranea* Seurat reference object
- identifies neoblast cells using a configurable metadata column and regular expression
- subsets the Seurat object to neoblast cells
- rebuilds a clean RNA-counts-based Seurat object for export
- writes the neoblast subset as `.h5ad`
- optionally saves the neoblast-only Seurat object as `.rds`

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(zellkonverter)
library(SingleCellExperiment)
library(knitr)
```

### Parameters

```{r print_parameters, echo=TRUE}
str(params)
```

## Resolve paths

```{r resolve_paths, echo=TRUE}
output_dir <- dirname(params$output_h5ad_file)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

path_table <- data.frame(
  object = c(
    "Input annotated Seurat object",
    "Output neoblast H5AD object",
    "Output neoblast Seurat object"
  ),
  path = c(
    params$input_seurat_file,
    params$output_h5ad_file,
    params$output_seurat_file
  ),
  exists = file.exists(c(
    params$input_seurat_file,
    params$output_h5ad_file,
    params$output_seurat_file
  )),
  stringsAsFactors = FALSE
)

kable(path_table, caption = "Input and output paths used by this report")
```

## Load annotated reference object

```{r load_reference_object, echo=TRUE}
if (!file.exists(params$input_seurat_file)) {
  stop("Input Seurat object not found: ", params$input_seurat_file)
}

srt_obj <- readRDS(params$input_seurat_file)

if (!inherits(srt_obj, "Seurat")) {
  stop("The input file did not contain a Seurat object.")
}

if (!params$annotation_col %in% colnames(srt_obj@meta.data)) {
  stop(
    "annotation_col not found in Seurat metadata: ",
    params$annotation_col,
    "\nAvailable columns include:\n",
    paste(colnames(srt_obj@meta.data), collapse = ", ")
  )
}

assay_names <- names(srt_obj@assays)

if (is.null(params$assay) || !is.character(params$assay) || length(params$assay) != 1) {
  stop("params$assay must be a single character value, e.g. assay: 'RNA'")
}

if (!params$assay %in% assay_names) {
  stop(
    "Requested assay not found in object: ", params$assay,
    "\nAvailable assays:\n",
    paste(assay_names, collapse = ", ")
  )
}

DefaultAssay(srt_obj) <- params$assay

reference_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay", "annotation_col"),
  value = c(nrow(srt_obj), ncol(srt_obj), DefaultAssay(srt_obj), params$annotation_col),
  stringsAsFactors = FALSE
)

kable(reference_summary, caption = "Summary of the input annotated reference object")
```

## Subset neoblast cells

```{r identify_cells, echo=TRUE}
clusters_to_keep <- unlist(params$clusters_to_keep)

cells_to_keep <- rownames(srt_obj@meta.data)[
  srt_obj@meta.data[[params$annotation_col]] %in% clusters_to_keep
]

if (length(cells_to_keep) == 0) {
  stop("No cells matched the requested cluster IDs.")
}

srt_obj_subset <- subset(
  x = srt_obj,
  cells = cells_to_keep
)

subset_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay"),
  value = c(nrow(srt_obj_subset), ncol(srt_obj_subset), DefaultAssay(srt_obj_subset)),
  stringsAsFactors = FALSE
)

kable(subset_summary, caption = "Summary of the neoblast-only Seurat object")

```

## Rebuild counts-based export object

```{r rebuild_counts_based_object, echo=TRUE}

srt_obj_subset <- JoinLayers(srt_obj_subset)

rna_counts <- Seurat::GetAssayData(
  object = srt_obj_subset,
  assay = params$assay,
  layer = params$layer
)

srt_obj_subset_export <- Seurat::CreateSeuratObject(
  counts = rna_counts,
  meta.data = srt_obj_subset@meta.data,
  project = "S_mediterranea_neoblast"
)

new_col_name <- params$final_label_col
cluster_id_col <- params$cluster_id_col
root_cluster_id <- params$root_cluster_id

final_label <- srt_obj_subset_export[[cluster_id_col]][,1]

final_label <- paste0(root_cluster_id, final_label)

srt_obj_subset_export[[new_col_name]] <- final_label

export_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay"),
  value = c(nrow(srt_obj_subset_export), ncol(srt_obj_subset_export), DefaultAssay(srt_obj_subset_export)),
  stringsAsFactors = FALSE
)

kable(export_summary, caption = "Summary of the counts-based object prepared for H5AD export")
```

## Export neoblast subset

```{r export_subset, echo=TRUE}
sce <- Seurat::as.SingleCellExperiment(srt_obj_subset_export)

zellkonverter::writeH5AD(
  sce = sce,
  file = params$output_h5ad_file
)

if (isTRUE(params$save_intermediate_files)) {
  saveRDS(srt_obj_subset, params$output_seurat_file)
}

output_table <- data.frame(
  object = c("S. mediterranea neoblast H5AD object", "S. mediterranea neoblast Seurat object"),
  file = c(params$output_h5ad_file, params$output_seurat_file),
  exists = file.exists(c(params$output_h5ad_file, params$output_seurat_file)),
  stringsAsFactors = FALSE
)

kable(output_table, caption = "Output files produced by this report")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
