---
title: "07_PrepareReferenceAtlasSMediterranea"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"
  basedir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mediterranea"
  principal_matrix_file: "GSE111764_PrincipalClusteringDigitalExpressionMatrix.dge.txt.gz"
  brain_matrix_file: "GSE111764_BrainClusteringDigitalExpressionMatrix.dge.txt.gz"
  sexual_matrix_file: "GSE111764_SexualClusteringDigitalExpressionMatrix.dge.txt.gz"
  cell_annotation_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mediterranea/S_mediterranea_metadata.txt"
  marker_gene_file: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas/data/metadata/S_med_marker_genes.tsv"
  include_sexual: FALSE
  output_object_file: "S_mediterranea_seurat.rds"
  output_annotated_object_file: "S_mediterranea_seurat_annotated.rds"
  save_intermediate_files: TRUE
  save_figures: TRUE
  figure_dpi: 600
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mediterranea/"
  seed: 42
  min_cells: 3
  min_features: 200
  npcs: 100
  integration_dims: 40
  clustering_resolution: 0.6
  k_param: 20
---

# Description

Create a Seurat reference atlas for *Schmidtea mediterranea* from the digital gene expression
matrices provided by the published dataset.

This report:

- reads the Principal and Brain DGE matrices, and optionally the Sexual matrix
- creates Seurat objects and attaches sample metadata
- normalizes the RNA assay and computes cell cycle scores
- merges the selected reference objects
- performs SCTransform, PCA, layer integration, UMAP, neighbor graph construction, and clustering
- attaches external cell type annotations
- generates reference plots and marker-based diagnostic figures

The resulting object is intended to serve as a reference atlas for cross-species cell type annotation.

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
library(dplyr)
library(harmony)
library(scCustomize)
library(speckle)
library(scater)
library(DoubletFinder)
library(knitr)
library(DT)
library(patchwork)
library(ComplexHeatmap)
```

### Parameters

```{r print_parameters, echo=TRUE}
str(params)
```

### Load helper functions

```{r load_functions, echo=TRUE}
bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "seurat.R",
    "smed_reference_helpers.R"
  )
)
```

### Prepare output directories

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)

figure_dir <- file.path(params$out_dir, "figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

obj_dir <- file.path(params$out_dir, "objs")
dir.create(obj_dir, recursive = TRUE, showWarnings = FALSE)

cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n\n")

cat("Figure directory:\n")
cat(normalizePath(figure_dir, mustWork = TRUE), "\n\n")

cat("Object directory:\n")
cat(normalizePath(obj_dir, mustWork = TRUE), "\n")
```

## Resolve input files

```{r resolve_input_files, echo=TRUE}
principal_path <- file.path(params$basedir, params$principal_matrix_file)
brain_path <- file.path(params$basedir, params$brain_matrix_file)
sexual_path <- file.path(params$basedir, params$sexual_matrix_file)
annotation_path <- file.path(params$basedir, params$cell_annotation_file)

input_files <- data.frame(
  input = c("Principal matrix", "Brain matrix", "Sexual matrix", "Cell annotation"),
  path = c(principal_path, brain_path, sexual_path, annotation_path),
  exists = file.exists(c(principal_path, brain_path, sexual_path, annotation_path)),
  stringsAsFactors = FALSE
)

kable(input_files, caption = "Input files used by this report")
```

## Define cell cycle features

```{r define_cell_cycle_features, echo=TRUE}
cell_cycle_features <- get_smed_cell_cycle_features()

kable(
  data.frame(
    feature_set = c("S", "G2M"),
    n_genes = c(length(cell_cycle_features$s_features), length(cell_cycle_features$g2m_features))
  ),
  caption = "Number of S. mediterranea cell cycle genes used for scoring"
)
```

## Read DGE matrices

```{r read_dge_matrices, echo=TRUE}
dge_list <- read_smed_dge_matrices(
  principal_path = principal_path,
  brain_path = brain_path,
  sexual_path = sexual_path,
  include_sexual = isTRUE(params$include_sexual)
)

dge_summary <- bind_rows(lapply(names(dge_list), function(nm) {
  mat <- dge_list[[nm]]
  data.frame(
    sample = nm,
    n_features = nrow(mat),
    n_cells = ncol(mat),
    stringsAsFactors = FALSE
  )
}))

kable(dge_summary, caption = "Summary of imported DGE matrices")
```

## Create Seurat objects

```{r create_seurat_objects, echo=TRUE}
seurat_list <- create_smed_seurat_list(
  dge_list = dge_list,
  min.cells = params$min_cells,
  min.features = params$min_features
)

seurat_summary <- bind_rows(lapply(names(seurat_list), function(nm) {
  obj <- seurat_list[[nm]]
  data.frame(
    sample = nm,
    n_features = nrow(obj),
    n_cells = ncol(obj),
    sample_ID = unique(obj$sample_ID),
    stringsAsFactors = FALSE
  )
}))

kable(seurat_summary, caption = "Summary of created Seurat objects")
```

## Normalize RNA and compute cell cycle scores

```{r normalize_and_score_cell_cycle, echo=TRUE}
seurat_list_scored <- lapply(seurat_list, function(obj) {
  normalize_and_score_smed_cell_cycle(
    obj = obj,
    s_features = cell_cycle_features$s_features,
    g2m_features = cell_cycle_features$g2m_features
  )
})

cell_cycle_summary <- bind_rows(lapply(names(seurat_list_scored), function(nm) {
  obj <- seurat_list_scored[[nm]]
  data.frame(
    sample = nm,
    mean_S_Score = round(mean(obj$S.Score, na.rm = TRUE), 4),
    mean_G2M_Score = round(mean(obj$G2M.Score, na.rm = TRUE), 4),
    stringsAsFactors = FALSE
  )
}))

kable(cell_cycle_summary, caption = "Cell cycle score summary for each sample")
```

## Merge reference objects

```{r merge_reference_objects, echo=TRUE}
reference_obj <- merge_smed_reference_objects(
  obj_list = seurat_list_scored,
  project = "scRNA_Seq_S_mediterranea"
)

cat("Merged reference dimensions (features x cells):\n")
print(dim(reference_obj))
```

## Build or load integrated reference

```{r integrate_reference, echo=TRUE}
reference_file <- file.path(obj_dir, params$output_object_file)

if (file.exists(reference_file)) {
  srt_obj_raw <- readRDS(reference_file)
  cat("Reusing cached integrated reference object:\n")
  cat("  ", reference_file, "\n")
} else {
  options(future.globals.maxSize = 50000 * 1024^2)

  srt_obj_raw <- run_smed_reference_workflow(
    obj = reference_obj,
    seed = params$seed,
    npcs = params$npcs,
    integration_dims = seq_len(params$integration_dims),
    clustering_resolution = params$clustering_resolution,
    k_param = params$k_param
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(srt_obj_raw, reference_file)
  }
}

cat("Integrated reference dimensions (features x cells):\n")
print(dim(srt_obj_raw))
```

## Attach cell type annotations

```{r attach_celltype_annotations, echo=TRUE}
annotated_reference_file <- file.path(params$out_dir, params$output_annotated_object_file)

if (file.exists(annotated_reference_file)) {
  srt_obj <- readRDS(annotated_reference_file)
  cat("Reusing cached annotated reference object:\n")
  cat("  ", annotated_reference_file, "\n")
} else {
  cell_metadata <- read_smed_cell_annotation_table(params$cell_annotation_file)

  srt_obj <- attach_smed_cell_labels(
    obj = srt_obj_raw,
    cell_metadata = cell_metadata,
    cell_id_col = "Cell.ID",
    metadata_cols = c("Major.cluster.description", "Cluster.number", "Cluster.ID", "Subcluster.Number", "Subcluster.ID"),
    strip_prefix_regex = "^[^_]*_"
  )

  if (isTRUE(params$save_intermediate_files)) {
    saveRDS(srt_obj, annotated_reference_file)
  }
}

annotation_summary <- data.frame(
  metric = c("n_cells", "n_features", "n_celltype_labels_non_missing"),
  value = c(
    ncol(srt_obj),
    nrow(srt_obj),
    sum(!is.na(srt_obj$Major.cluster.description))
  ),
  stringsAsFactors = FALSE
)

kable(annotation_summary, caption = "Summary after adding cell type annotations")
```

## Reference overview plots

```{r reference_overview_plots, echo=TRUE, fig.width=12, fig.height=6}
p_dim_default <- scCustomize::DimPlot_scCustom(
  srt_obj_raw,
  aspect_ratio = 1
)

p_dim_celltype <- scCustomize::DimPlot_scCustom(
  srt_obj,
  aspect_ratio = 1,
  group.by = "Major.cluster.description"
)

p_qc <- scCustomize::FeaturePlot_scCustom(
  seurat_object = srt_obj_raw,
  aspect_ratio = 1,
  features = c("nCount_RNA", "nFeature_RNA"),
  pt.size = 0.1,
  num_columns = 1
)

print(p_dim_default)
print(p_dim_celltype)
print(p_qc)
```

## Marker-based feature plots

```{r marker_feature_plots, echo=TRUE, fig.width=12, fig.height=8}
marker_table <- read_marker_dotplot_table(params$marker_gene_file)

marker_feature_table <- build_marker_feature_plot_table(marker_table)

kable(
  marker_feature_table,
  caption = "Marker genes used for reference feature plots"
)

marker_plot_list <- lapply(seq_len(nrow(marker_feature_table)), function(i) {
  marker_name <- marker_feature_table$CellType[i]
  gene_id <- marker_feature_table$GeneID[i]

  p <- plot_smed_marker_feature(
    obj = srt_obj_raw,
    feature = gene_id,
    title = paste0(marker_name, ": ", gene_id)
  )

  print(p)
  p
})

names(marker_plot_list) <- paste(
  marker_feature_table$CellType,
  marker_feature_table$GeneID,
  sep = " | "
)
```

## Marker dot plot

```{r marker_dotplot, echo=TRUE, fig.width=12, fig.height=8}
kable(
  marker_table,
  caption = "Marker genes used for the reference dot plot"
)

marker_info <- build_marker_dotplot_features(marker_table)

p_dot <- plot_smed_marker_dotplot(
  obj = srt_obj,
  features = marker_info$features,
  group_by = "Major.cluster.description"
)

print(p_dot)
```

## Save publication-ready figures

```{r save_figures, echo=TRUE}
if (isTRUE(params$save_figures)) {
  save_smed_reference_plot(
    plot_obj = p_dim_default,
    file_stub = "dimplot_reference_default",
    out_dir = figure_dir,
    dpi = params$figure_dpi
  )

  save_smed_reference_plot(
    plot_obj = p_dim_celltype,
    file_stub = "dimplot_reference_celltype",
    out_dir = figure_dir,
    dpi = params$figure_dpi
  )

  save_smed_reference_plot(
    plot_obj = p_qc,
    file_stub = "featureplot_reference_qc",
    out_dir = figure_dir,
    dpi = params$figure_dpi
  )

  for (nm in names(marker_plot_list)) {
    save_smed_reference_plot(
      plot_obj = marker_plot_list[[nm]],
      file_stub = paste0("featureplot_marker_", gsub("[^A-Za-z0-9]+", "_", tolower(nm))),
      out_dir = figure_dir,
      dpi = params$figure_dpi
    )
  }

  save_smed_reference_plot(
    plot_obj = p_dot,
    file_stub = "dotplot_reference_markers",
    out_dir = figure_dir,
    width = 12,
    height = 8,
    dpi = params$figure_dpi
  )
}
```

## Output files

```{r output_files_summary, echo=TRUE}
output_files <- data.frame(
  object = c(
    "Integrated S. mediterranea reference object",
    "Annotated S. mediterranea reference object"
  ),
  file = c(
    reference_file,
    annotated_reference_file
  ),
  exists = file.exists(c(reference_file, annotated_reference_file)),
  stringsAsFactors = FALSE
)

kable(output_files, caption = "Objects produced by this report")
```

```{r figure_file_summary, echo=TRUE}
saved_figures <- list.files(figure_dir, full.names = FALSE)

figure_table <- data.frame(
  figure_file = saved_figures,
  stringsAsFactors = FALSE
)

kable(figure_table, caption = "Publication-ready figures saved to disk")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
