---
title: "08_2_Subset_S_mansoni_Parenchyma_to_H5AD"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mansoni/" 
  input_seurat_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mansoni/S_mansoni_seurat_annotated.rds"
  output_h5ad_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mansoni/S_mansoni_parenchyma_cells.h5ad"
  output_seurat_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/data/ref_atlas/S_mansoni/S_mansoni_parenchyma_cells.rds"
  annotation_col: "Cluster.number"
  clusters_to_keep:
    value:
      - "6"
      - "19"
      - "30"
      - "43"
  assay: "RNA"
  layer: "counts"
  save_intermediate_files: TRUE
  seed: 42
---
  
# Description
  
Convert the schistosome reference atlas from its original Seurat object into an updated Seurat v5-compatible object, add broad cell type annotations, visualize the resulting labels.

This report:
  
- loads the original Seurat object
- removes stale integrated `scale.data` to allow object upgrading
- upgrades the object structure with `UpdateSeuratObject()`
- derives `cell_type` and `broad_cell_type` metadata
- visualizes the broad cell type annotations
- rebuilds a lean RNA-counts-based Seurat object for export
- saves the upgraded Seurat v5 object as `.rds`

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(scCustomize)
library(ggplot2)
library(knitr)
```

### Parameters

```{r print_parameters, echo=TRUE}
str(params)
```

### Prepare output directory

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)

cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n")
```

## Resolve input and output paths

```{r resolve_paths, echo=TRUE}
input_seurat_path <- file.path(params$working_dir, params$input_seurat_file)
output_seurat_v5_path <- file.path(params$working_dir, params$output_seurat_v5_file)

path_table <- data.frame(
  object = c(
    "Input Seurat object",
    "Output Seurat v5 object"
  ),
  path = c(
    input_seurat_path,
    output_seurat_v5_path
  ),
  exists = c(
    file.exists(input_seurat_path),
    file.exists(output_seurat_v5_path)
  ),
  stringsAsFactors = FALSE
)

kable(path_table, caption = "Input and output paths used by this report")
```

## Load original Seurat object

```{r load_seurat_object, echo=TRUE}
if (!file.exists(input_seurat_path)) {
  stop("Input Seurat object not found: ", input_seurat_path)
}

srt_obj <- readRDS(input_seurat_path)

```

## Repair object structure and update to Seurat v5

```{r update_seurat_object, echo=TRUE}
old_opt <- getOption("Seurat.object.validate", TRUE)

options(Seurat.object.validate = FALSE)

if ("integrated" %in% Assays(srt_obj)) {
  srt_obj@assays[["integrated"]]@scale.data <- matrix(numeric(0), nrow = 0, ncol = 0)
}

srt_obj <- UpdateSeuratObject(srt_obj)

options(Seurat.object.validate = old_opt)

obj_summary_after <- data.frame(
  metric = c("n_features", "n_cells", "default_assay"),
  value = c(
    nrow(srt_obj),
    ncol(srt_obj),
    DefaultAssay(srt_obj)
  ),
  stringsAsFactors = FALSE
)

kable(obj_summary_after, caption = "Summary after updating the Seurat object")
```

## Add broad cell type metadata

```{r add_celltype_metadata, echo=TRUE}
srt_obj$cell_type <- Idents(srt_obj)
srt_obj$broad_cell_type <- ""

srt_obj$broad_cell_type <- ifelse(grepl(pattern = "neuron", x = srt_obj$cell_type), "neuron", "")
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "muscle", x = srt_obj$cell_type), "muscle", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "flame cells", x = srt_obj$cell_type), "flame cells", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "parenchyma", x = srt_obj$cell_type), "parenchyma", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "zfp-1-1+", x = srt_obj$cell_type), "tegument lineage", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "egc+", x = srt_obj$cell_type), "tegument lineage", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "meg-1+", x = srt_obj$cell_type), "tegument lineage", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "early tsp-2+", x = srt_obj$cell_type), "tegument lineage", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "sm13+", x = srt_obj$cell_type), "tegument lineage", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "tegument 1", x = srt_obj$cell_type), "tegument lineage", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "GSCs", x = srt_obj$cell_type), "germline", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "GSC progeny", x = srt_obj$cell_type), "germline", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "female gametes", x = srt_obj$cell_type), "female gamete", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "^male gametes", x = srt_obj$cell_type), "male gamete", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "gut", x = srt_obj$cell_type), "gut", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "oesophageal gland", x = srt_obj$cell_type), "esophageal gland", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "neoblast", x = srt_obj$cell_type), "neoblast", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse((srt_obj$cell_type == "neoblast progeny"), "neoblast progeny", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "vitellocytes", x = srt_obj$cell_type), "vittelaria", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "S1 progeny", x = srt_obj$cell_type), "vittelaria", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse((srt_obj$cell_type == "S1"), "vittelaria", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse((srt_obj$cell_type == "prom2+"), "prom2+", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "hes2+", x = srt_obj$cell_type), "hes2+", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse(grepl(pattern = "Mehlis' gland", x = srt_obj$cell_type), "Mehlis' gland", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse((srt_obj$cell_type == "tegument 2"), "tegument 2", srt_obj$broad_cell_type)
srt_obj$broad_cell_type <- ifelse((srt_obj$cell_type == "unknown"), "unknown", srt_obj$broad_cell_type)

celltype_summary <- as.data.frame(table(srt_obj$broad_cell_type), stringsAsFactors = FALSE)
colnames(celltype_summary) <- c("broad_cell_type", "n_cells")

kable(celltype_summary, caption = "Broad cell type annotation summary")
```

## Visualize broad cell types

```{r plot_broad_cell_types, echo=TRUE, fig.width=10, fig.height=8}
DefaultAssay(srt_obj) <- "RNA"

p_broad <- DimPlot_scCustom(
  srt_obj,
  group.by = "broad_cell_type",
  aspect_ratio = 1
)

print(p_broad)
```

## Rebuild export object from RNA counts

```{r rebuild_export_object, echo=TRUE}
rna_counts <- GetAssayData(srt_obj, assay = "RNA", layer = "counts")

new_srt <- CreateSeuratObject(counts = rna_counts)

original_metadata <- srt_obj@meta.data
new_srt <- AddMetaData(new_srt, metadata = original_metadata)

export_summary <- data.frame(
  metric = c("n_features", "n_cells", "default_assay"),
  value = c(
    nrow(new_srt),
    ncol(new_srt),
    DefaultAssay(new_srt)
  ),
  stringsAsFactors = FALSE
)

kable(export_summary, caption = "Summary of the exported counts-based Seurat object")
```

## Export to H5AD and Seurat RDS

```{r export_objects, echo=TRUE}
saveRDS(srt_obj, output_seurat_v5_path)

output_table <- data.frame(
  object = c("Seurat v5 RDS export"),
  file = c(output_seurat_v5_path),
  exists = file.exists(output_seurat_v5_path),
  stringsAsFactors = FALSE
)

kable(output_table, caption = "Exported reference objects")
```

## Save figure

```{r save_figure, echo=TRUE}
if (isTRUE(params$save_figures)) {
  ggplot2::ggsave(
    filename = file.path(params$out_dir, "dimplot_broad_cell_type_schistosome.pdf"),
    plot = p_broad,
    width = 10,
    height = 8,
    units = "in",
    dpi = params$figure_dpi,
    bg = "white"
  )
  
  ggplot2::ggsave(
    filename = file.path(params$out_dir, "dimplot_broad_cell_type_schistosome.png"),
    plot = p_broad,
    width = 10,
    height = 8,
    units = "in",
    dpi = params$figure_dpi,
    bg = "white"
  )
}
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
