---
title: "09_1_GO_Enrichment_Gut"
author: "Francisco Pereira Lobo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
    code_folding: "hide"
    fig_caption: true
params:
  project_root: "~/Projects/Erin/M_lignano/Mlig_scRNASeq_atlas"

  input_seurat_file: "//data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/objs/Digestive_System_integrated_merged.rds"
  gene2go_file: "/data/Mlig_scRNA_Seq/Francisco/data/geneid2go.map"

  markers_file: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results/tables/FindAllMarkers_for_GO.tsv"
  recompute_markers: FALSE

  marker_assay: "RNA"
  marker_ident_col: "digestive_cluster"
  only_pos_markers: TRUE
  findallmarkers_logfc_threshold: 0
  findallmarkers_min_pct: 0.1
  findallmarkers_test_use: "wilcox"

  marker_cluster_col: "cluster"
  marker_gene_col: "gene"
  marker_padj_col: "p_val_adj"
  marker_logfc_col: "avg_log2FC"

  ontology: "BP"
  padj_threshold: 0.01
  logfc_threshold: 0.5
  topgo_fdr_threshold: 0.2
  node_size: 3
  top_n_terms_per_cluster: 200

  save_intermediate_files: TRUE
  save_figures: TRUE
  figure_dpi: 600
  out_dir: "/data/Mlig_scRNA_Seq/Francisco/Mlig_scRNASeq_atlas/results"
  seed: 42
---

# Description

Compute Gene Ontology enrichment from marker genes generated with `Seurat::FindAllMarkers()`.

This report:

- loads an integrated Seurat object
- sets cell identities from a metadata column
- computes or reuses a `FindAllMarkers()` table
- filters marker genes by adjusted p-value and log fold-change
- runs topGO enrichment independently for each cluster or cell type
- saves all and significant enrichment results
- generates a gprofiler-style dot plot of enriched GO terms grouped by GO ontology category

## RMarkdown Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 12,
  fig.height = 8,
  dpi = 150
)

set.seed(params$seed)
```

## Preparing the environment

### Loading libraries

```{r load_libraries, echo=TRUE}
library(Seurat)
library(dplyr)
library(stringr)
library(ggplot2)
library(forcats)
library(DT)
library(knitr)
library(topGO)
library(gprofiler2)
```

### Parameters

```{r print_parameters, echo=TRUE}
str(params)
```

### Load helper functions

```{r load_functions, echo=TRUE}
bootstrap <- file.path(
  normalizePath(params$project_root, mustWork = TRUE),
  "R",
  "source_report_functions.R"
)

if (!file.exists(bootstrap)) {
  stop("Bootstrap file not found at: ", bootstrap)
}

source(bootstrap)

source_report_functions(
  base_dir = params$project_root,
  rel_dir  = "R/report_utils",
  files    = c(
    "io.R",
    "go_enrichment_helpers.R"
  )
)
```

### Prepare output directories

```{r prepare_output_dir, echo=TRUE}
dir.create(params$out_dir, recursive = TRUE, showWarnings = FALSE)

table_dir <- file.path(params$out_dir, "tables")
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)

figure_dir <- file.path(params$out_dir, "figures", "go_enrichment")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

dir.create(dirname(params$markers_file), recursive = TRUE, showWarnings = FALSE)

cat("Output directory:\n")
cat(normalizePath(params$out_dir, mustWork = TRUE), "\n\n")
cat("Table directory:\n")
cat(normalizePath(table_dir, mustWork = TRUE), "\n\n")
cat("Figure directory:\n")
cat(normalizePath(figure_dir, mustWork = TRUE), "\n\n")
cat("FindAllMarkers file:\n")
cat(params$markers_file, "\n")
```

## Resolve input files

```{r resolve_input_files, echo=TRUE}
input_files <- data.frame(
  input = c("Seurat object", "gene2GO map"),
  path = c(params$input_seurat_file, params$gene2go_file),
  exists = file.exists(c(params$input_seurat_file, params$gene2go_file)),
  stringsAsFactors = FALSE
)

kable(input_files, caption = "Required input files used by this report")

if (!all(input_files$exists)) {
  stop("One or more required input files are missing.")
}
```

## Load Seurat object

```{r load_seurat_object, echo=TRUE}
srt_obj <- readRDS(params$input_seurat_file)

DefaultAssay(srt_obj) <- params$marker_assay

srt_obj <- JoinLayers(srt_obj)

if (!inherits(srt_obj, "Seurat")) {
  stop("input_seurat_file did not contain a Seurat object.")
}

if (!params$marker_ident_col %in% colnames(srt_obj@meta.data)) {
  stop(
    "marker_ident_col not found in Seurat metadata: ", params$marker_ident_col,
    "\nAvailable columns:\n",
    paste(colnames(srt_obj@meta.data), collapse = ", ")
  )
}

available_assays <- names(srt_obj@assays)

if (!params$marker_assay %in% available_assays) {
  stop(
    "marker_assay not found in object: ", params$marker_assay,
    "\nAvailable assays:\n",
    paste(available_assays, collapse = ", ")
  )
}

input_summary <- data.frame(
  metric = c(
    "n_cells",
    "n_features",
    "default_assay_before_marker_testing",
    "marker_assay",
    "marker_identity_column",
    "n_identities"
  ),
  value = c(
    ncol(srt_obj),
    nrow(srt_obj),
    DefaultAssay(srt_obj),
    params$marker_assay,
    params$marker_ident_col,
    length(unique(srt_obj@meta.data[[params$marker_ident_col]]))
  ),
  stringsAsFactors = FALSE
)

kable(input_summary, caption = "Seurat object summary before marker testing")
```

## Compute or load FindAllMarkers table

```{r compute_or_load_markers, echo=TRUE}
markers <- compute_or_load_findallmarkers(
  obj = srt_obj,
  markers_file = params$markers_file,
  recompute_markers = params$recompute_markers,
  assay = params$marker_assay,
  ident_col = params$marker_ident_col,
  only_pos = params$only_pos_markers,
  logfc_threshold = params$findallmarkers_logfc_threshold,
  min_pct = params$findallmarkers_min_pct,
  test_use = params$findallmarkers_test_use,
  cluster_col = params$marker_cluster_col,
  gene_col = params$marker_gene_col,
  padj_col = params$marker_padj_col,
  logfc_col = params$marker_logfc_col
)

marker_summary <- data.frame(
  metric = c(
    "n_marker_rows",
    "n_unique_marker_genes",
    "n_marker_clusters"
  ),
  value = c(
    nrow(markers),
    length(unique(markers[[params$marker_gene_col]])),
    length(unique(markers[[params$marker_cluster_col]]))
  ),
  stringsAsFactors = FALSE
)

kable(marker_summary, caption = "FindAllMarkers output summary")

datatable(
  head(markers, 100),
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "First 100 rows of FindAllMarkers output"
)
```

## Load gene2GO annotation

```{r load_gene2go, echo=TRUE}
gene2GO <- read_gene2go_map(params$gene2go_file)

gene2go_summary <- data.frame(
  metric = c("n_annotated_genes", "n_go_assignments"),
  value = c(
    length(gene2GO),
    sum(vapply(gene2GO, length, integer(1)))
  ),
  stringsAsFactors = FALSE
)

kable(gene2go_summary, caption = "gene2GO annotation summary")
```

## Define topGO gene universe

```{r define_gene_universe, echo=TRUE}
all_genes <- rownames(srt_obj)
all_genes <- intersect(all_genes, names(gene2GO))

if (length(all_genes) == 0) {
  stop("No overlap between Seurat object genes and gene2GO annotation.")
}

universe_summary <- data.frame(
  metric = c(
    "n_genes_in_seurat_object",
    "n_genes_in_gene2GO",
    "n_genes_in_topGO_universe"
  ),
  value = c(
    nrow(srt_obj),
    length(gene2GO),
    length(all_genes)
  ),
  stringsAsFactors = FALSE
)

kable(universe_summary, caption = "topGO gene universe summary")
```

## Marker filtering summary

```{r marker_filtering_summary, echo=TRUE}
marker_filter_summary <- summarize_marker_selection_for_go(
  markers = markers,
  cluster_col = params$marker_cluster_col,
  gene_col = params$marker_gene_col,
  padj_col = params$marker_padj_col,
  logfc_col = params$marker_logfc_col,
  padj_threshold = params$padj_threshold,
  logfc_threshold = params$logfc_threshold
)

datatable(
  marker_filter_summary,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Marker genes retained for GO enrichment after filtering"
)
```

## Save selected marker genes used for GO enrichment

```{r save_selected_marker_genes, echo=TRUE}
selected_marker_genes <- get_selected_marker_genes_for_go(
  markers = markers,
  cluster_col = params$marker_cluster_col,
  gene_col = params$marker_gene_col,
  padj_col = params$marker_padj_col,
  logfc_col = params$marker_logfc_col,
  padj_threshold = params$padj_threshold,
  logfc_threshold = params$logfc_threshold,
  all_genes = all_genes
)

selected_marker_file <- file.path(
  table_dir,
  paste0("FindAllMarkers_selected_genes_for_topGO_", params$ontology, ".tsv")
)

if (isTRUE(params$save_intermediate_files)) {
  write.table(
    selected_marker_genes,
    file = selected_marker_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}

datatable(
  selected_marker_genes,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Selected marker genes used as foreground sets for topGO"
)
```

## Run topGO enrichment by cluster

```{r run_topgo_enrichment, echo=TRUE}
results_enrichment <- run_topgo_enrichment_by_cluster(
  markers = markers,
  all_genes = all_genes,
  gene2GO = gene2GO,
  cluster_col = params$marker_cluster_col,
  gene_col = params$marker_gene_col,
  padj_col = params$marker_padj_col,
  logfc_col = params$marker_logfc_col,
  padj_threshold = params$padj_threshold,
  logfc_threshold = params$logfc_threshold,
  ontology = params$ontology,
  node_size = params$node_size
)

results_file <- file.path(
  table_dir,
  paste0("topGO_enrichment_", params$ontology, "_FindAllMarkers.tsv")
)

if (isTRUE(params$save_intermediate_files)) {
  write.table(
    results_enrichment,
    file = results_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}

datatable(
  results_enrichment,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "All topGO enrichment results"
)
```

## Significant enrichment results

```{r significant_enrichment_results, echo=TRUE}
sig_results <- results_enrichment %>%
  dplyr::filter(
    !is.na(.data$Fisher_FDR),
    .data$Fisher_FDR < params$topgo_fdr_threshold
  ) %>%
  dplyr::arrange(.data$ClusterID, .data$Fisher_FDR)

sig_results_file <- file.path(
  table_dir,
  paste0("topGO_enrichment_", params$ontology, "_FindAllMarkers_significant.tsv")
)

if (isTRUE(params$save_intermediate_files)) {
  write.table(
    sig_results,
    file = sig_results_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}

sig_summary <- sig_results %>%
  dplyr::count(.data$ClusterID, name = "n_significant_terms") %>%
  dplyr::arrange(dplyr::desc(.data$n_significant_terms))

kable(sig_summary, caption = "Number of significant enriched GO terms per cluster")

datatable(
  sig_results,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "Significant topGO enrichment results"
)
```

## Prepare gprofiler-style GO dot plot table

```{r prepare_gostplot_table, echo=TRUE}
gost_style_results <- topgo_to_gostplot_table(
  results = sig_results,
  ontology = params$ontology,
  max_terms_per_cluster = params$top_n_terms_per_cluster
)

gost_style_file <- file.path(
  table_dir,
  paste0("topGO_enrichment_", params$ontology, "_gostplot_style.tsv")
)

if (isTRUE(params$save_intermediate_files)) {
  write.table(
    gost_style_results,
    file = gost_style_file,
    sep = "\t",
    quote = FALSE,
    row.names = FALSE
  )
}

datatable(
  gost_style_results,
  options = list(pageLength = 20, scrollX = TRUE),
  caption = "topGO results converted to gprofiler-style format"
)
```

## GO enrichment dot plot

```{r go_dotplot, echo=TRUE, fig.width=14, fig.height=10}
p_go_dot <- plot_go_enrichment_dotplot(
  gost_style_results,
  top_n_terms_per_cluster = params$top_n_terms_per_cluster
)

print(p_go_dot)

if (isTRUE(params$save_figures)) {
  ggsave(
    filename = file.path(figure_dir, paste0("topGO_", params$ontology, "_dotplot.pdf")),
    plot = p_go_dot,
    width = 14,
    height = 10,
    units = "in",
    dpi = params$figure_dpi,
    bg = "white"
  )

  ggsave(
    filename = file.path(figure_dir, paste0("topGO_", params$ontology, "_dotplot.png")),
    plot = p_go_dot,
    width = 14,
    height = 10,
    units = "in",
    dpi = params$figure_dpi,
    bg = "white"
  )
}
```

## Optional gprofiler gostplot object

This section attempts to use `gprofiler2::gostplot()` on the topGO results converted into a gprofiler-like result table. If the conversion is not compatible with the installed `gprofiler2` version, this section fails gracefully and does not interrupt the report.

```{r optional_gostplot, echo=TRUE, fig.width=14, fig.height=8}
if (nrow(gost_style_results) > 0) {
  p_gost <- tryCatch(
    {
      gprofiler2::gostplot(
        gostres = list(result = gost_style_results),
        capped = TRUE,
        interactive = FALSE
      )
    },
    error = function(e) {
      message("gprofiler2::gostplot() could not plot converted topGO results: ", conditionMessage(e))
      NULL
    }
  )

  if (!is.null(p_gost)) {
    print(p_gost)

    if (isTRUE(params$save_figures)) {
      ggsave(
        filename = file.path(figure_dir, paste0("topGO_", params$ontology, "_gprofiler_gostplot.pdf")),
        plot = p_gost,
        width = 14,
        height = 8,
        units = "in",
        dpi = params$figure_dpi,
        bg = "white"
      )

      ggsave(
        filename = file.path(figure_dir, paste0("topGO_", params$ontology, "_gprofiler_gostplot.png")),
        plot = p_gost,
        width = 14,
        height = 8,
        units = "in",
        dpi = params$figure_dpi,
        bg = "white"
      )
    }
  }
}
```

## Output files

```{r output_files_summary, echo=TRUE}
output_files <- data.frame(
  object = c(
    "FindAllMarkers table",
    "Selected marker genes for topGO",
    "All topGO enrichment results",
    "Significant topGO enrichment results",
    "gprofiler-style topGO results",
    "GO dot plot PDF",
    "GO dot plot PNG"
  ),
  file = c(
    params$markers_file,
    selected_marker_file,
    results_file,
    sig_results_file,
    gost_style_file,
    file.path(figure_dir, paste0("topGO_", params$ontology, "_dotplot.pdf")),
    file.path(figure_dir, paste0("topGO_", params$ontology, "_dotplot.png"))
  ),
  exists = file.exists(c(
    params$markers_file,
    selected_marker_file,
    results_file,
    sig_results_file,
    gost_style_file,
    file.path(figure_dir, paste0("topGO_", params$ontology, "_dotplot.pdf")),
    file.path(figure_dir, paste0("topGO_", params$ontology, "_dotplot.png"))
  )),
  stringsAsFactors = FALSE
)

kable(output_files, caption = "Files produced by this report")
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
